---
title: "Urban Music Atlas"
format: 
  html:
    # Tabset layout chosen over format:dashboard to preserve custom CSS styling
    theme: darkly
    css: styles.css
    toc: false
    page-layout: full
---
```{r}
#| include: false
library(tidyverse)
library(compmus)
library(tidymodels)
library(ranger)
library(vip)
library(yardstick)
library(gt
        )

CM_corpus <- read_csv("canciones/CM_corpus.csv", show_col_types = FALSE)

CM_corpus <- CM_corpus |>
  rename_with(~ .x |>
    str_to_lower() |>
    str_replace_all("[^a-z0-9]+", "_") |>
    str_replace_all("^_|_$", ""))

DTMF <- read_csv('canciones/DTMF_chromagram.csv')
niBorracho <- read_csv('canciones/niBorracho_chromagram.csv')
malamente <- read_csv('canciones/malamente_chromagram.csv')

DTMF_timbre <- read_csv('canciones/DTMF_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_timbre <- read_csv('canciones/niBorracho_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
malamente_timbre <- read_csv('canciones/malamente_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_novelty <- read_csv("canciones/niBorracho_novelty.csv", show_col_types = FALSE)
niBorracho_act <- read_csv("canciones/niBorracho_act.csv", show_col_types = FALSE)
niBorracho_dft <- read_csv("canciones/niBorracho_dft.csv", show_col_types = FALSE)

# Shared plot theme
atlas_theme <- function() {
  theme_classic() +
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 10),
      axis.title       = element_text(color = "#8888aa", size = 11),
      plot.title       = element_text(color = "#f5c842", size = 16, face = "bold"),
      plot.subtitle    = element_text(color = "#f0ecdf", size = 11),
      plot.margin      = margin(12, 16, 12, 12),
      legend.background = element_rect(fill = "#12152a", color = NA),
      legend.key        = element_rect(fill = "#12152a", color = NA),
      legend.text       = element_text(color = "#f0ecdf"),
      legend.title      = element_text(color = "#f5c842")
    )
}

# Small helper to safely convert tempogram column names to numeric tempo bins
long_tempogram <- function(df) {
  time_col <- names(df)[1]

  df |>
    pivot_longer(-all_of(time_col), names_to = "tempo", values_to = "value") |>
    mutate(
      tempo = as.numeric(tempo)
    ) |>
    rename(TIME = all_of(time_col))
}
```

::: hero-banner
URBAN MUSIC ATLAS
:::

Spanish-language urban music (2015–2025) across three geographic contexts.

::: panel-tabset

## Background

<iframe src="https://open.spotify.com/embed/playlist/0Z93ne5Yrdqr1hLlRqF7hq" 
width="100%" height="380" frameborder="0" 
allowtransparency="true" allow="encrypted-media"
style="border-radius: 12px;"></iframe>

### Why this corpus?

Nowadays, there is a rise in Canarian reggaeton across Spain, and with 
it, a recurring debate in music criticism and on social media, *are artists like* 
*Quevedo genuinely expressing an Atlantic identity rooted in the Canary* 
*Islands, or are they simply mimicking Puerto Rican reggaeton?* This portfolio 
takes that question seriously and tries to answer it with data rather than 
opinion.

It is also worth acknowledging that **reggaeton is culturally Puerto Rican**. 
The genre was created by Puerto Rican and broader Caribbean communities, and 
any analysis of its spread must recognise that origin without flattening it into 
a generic "Latin sound." The goal here is not to claim equivalence between regions,
but to measure acoustic distance — and to let that distance speak.

> *Is Canarian urbano its own sound or is it just Bad Bunny with a Spanish accent?*

---

### The corpus

The corpus covers Spanish-language urban music from **2015 to 2025** across three 
geographic regions: **Puerto Rico**, the **Canary Islands**, and **Mainland Spain**.
Approximately **240 tracks** were assembled from representative artists in each 
region, balancing reggaeton-focused performers with hybrid urban artists. Tracks 
were retrieved via the Spotify API and selected to reflect each region's distinct
production ecosystem while keeping the corpus balanced enough for classification.

::: {.callout-note}
**Corpus assumption:** Artists were assigned to a single region based on their 
primary cultural origin, even when their music crosses regional lines. This is 
a deliberate simplification — collaborative tracks were assigned to the lead 
artist's region. The goal is to isolate regional signal, not to deny that 
these scenes are in constant dialogue with each other.
:::

---

### My three tracks

Three tracks were submitted to the class corpus — one per region — selected for 
their **cultural rootedness**, **sonic clarity**, and **popularity**. Each is 
emblematic of its region's production style, making them ideal anchors for 
cross-regional comparison.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DtMF · Bad Bunny** — Puerto Rico
:::
Bad Bunny is one of the most-streamed Latin artists of all time and the most 
prominent voice in global reggaeton. **DtMF** was chosen because it is deeply 
rooted in dembow rhythm, minor-key trap production, and Boricua street 
vernacular — a direct cultural export of Caribbean urbano. It represents 
the Puerto Rican end of the spectrum as clearly as any single track can.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands
:::
Quevedo became one of Spain's most-streamed artists with a sound heavily 
influenced by Latin Caribbean music. **NI BORRACHO** was chosen because it 
sits at the precise cultural crossroads this study investigates: rhythmically 
Caribbean, lyrically Canarian, produced in a style that


> *Are there measurable stylistic differences across regions?*

## Track Features

In particular tempo provides a useful rythmic dimension for comparing regional scenes.

```{r}
#| echo: false

ggplot(CM_corpus, aes(x = tempo)) +
  geom_histogram(
    binwidth = 5,
    fill = "#00b4a2",
    color = "#0a0c14"
  ) +
  labs(
    title = "Tempo distribution of the corpus",
    subtitle = "Spotify tempo estimates for the corpus",
    x = "Tempo (BPM)",
    y = "Number of tracks"
  ) +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    axis.title       = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold"),
    plot.subtitle    = element_text(color = "#f0ecdf")
  )
```

The histogram shows that most tracks cluster in a fairly tight rhythmic zone typical
of contemporary urbano and reggaeton. Much of the playlist lies roughly between
90 and 140 BPM, which is consistent with dembow-based rhythmic structures and
dance-oriented production. At the same time, the spread outside this central range
suggests hybridisation: some tracks slow down into heavier groove-based zones,
while others move upward toward faster pop-electronic tempos.

## Chroma Features

These three chromagrams represent the pitch-class energy over time for three 
culturally rooted tracks from each region.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DTMF · Bad Bunny** — Puerto Rico

:::
```{r}
#| echo: false
DTMF |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Shows prominent activity around A♭ and E, common in reggaeton production rooted 
in Puerto Rico.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands

:::
```{r}
#| echo: false
niBorracho |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Energy evenly spread across A, G, F, and E — no single dominant tonal centre, 
suggesting a blend of Iberian and Latin American influences.
:::

::: {.chroma-card .sp}
::: chroma-header
🇪🇸 **MALAMENTE · Rosalía** — Mainland Spain

:::
```{r}
#| echo: false
malamente |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Strong concentration around G, C, and F — reflecting the harmonic language deeply embedded in flamenco tradition.
:::

:::

> Quevedo's chromagram appears harmonically closer to Bad Bunny than to Rosalía — hinting that Canarian music carries a stronger Latin Caribbean influence than a Peninsular Spanish one.

---

### Key Estimation
```{r}
#| echo: false
#| warning: false
#| include: false
circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

major_key <- c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <- c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

key_templates <- tribble(
  ~name, ~template,
  "Gb:maj", circshift(major_key, 6),
  "Bb:min", circshift(minor_key, 10),
  "Db:maj", circshift(major_key, 1),
  "F:min",  circshift(minor_key, 5),
  "Ab:maj", circshift(major_key, 8),
  "C:min",  circshift(minor_key, 0),
  "Eb:maj", circshift(major_key, 3),
  "G:min",  circshift(minor_key, 7),
  "Bb:maj", circshift(major_key, 10),
  "D:min",  circshift(minor_key, 2),
  "F:maj",  circshift(major_key, 5),
  "A:min",  circshift(minor_key, 9),
  "C:maj",  circshift(major_key, 0),
  "E:min",  circshift(minor_key, 4),
  "G:maj",  circshift(major_key, 7),
  "B:min",  circshift(minor_key, 11),
  "D:maj",  circshift(major_key, 2),
  "F#:min", circshift(minor_key, 6),
  "A:maj",  circshift(major_key, 9),
  "C#:min", circshift(minor_key, 1),
  "E:maj",  circshift(major_key, 4),
  "G#:min", circshift(minor_key, 8),
  "B:maj",  circshift(major_key, 11),
  "D#:min", circshift(minor_key, 3)
)
```

These keygrams compare the chroma features of each track against major and minor
key profiles, revealing which key the music gravitates towards over time. 
Brighter areas indicate a stronger match to that key.

```{r}
#| echo: false
#| warning: false

# Shared theme helper — define once, reuse
keygram_theme <- function(title_text) {
  list(
    geom_tile(color = "#12152a", linewidth = 0.3),
    scale_fill_gradientn(
      colors = c("#d62839", "#0a0c14", "#00b4a2", "#f5c842"),
      guide  = "none"
    ),
    theme_classic(),
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 7),
      axis.text.y      = element_text(color = "#f0ecdf", size = 7),
      axis.title       = element_text(color = "#8888aa"),
      plot.title       = element_text(color = "#f5c842", family = "sans",
                                      size = 13, face = "bold",
                                      margin = margin(b = 10)),
      plot.margin      = margin(12, 16, 12, 12)
    ),
    labs(x = "Time (s)", y = "", title = title_text)
  )
}
```

**🇵🇷 DtMF · Bad Bunny — Keygram**
```{r}
#| echo: false
#| warning: false
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("DtMF — Keygram")
```

::: {.callout-note}
**Key:** The keygram shows consistent brightness concentrated around D#:min and 
B:maj throughout the track, suggesting DtMF gravitates towards these keys. 
This is consistent with reggaeton's tendency to use minor keys with a dark, 
intense character. The bright yellow spike in the first 25 seconds likely 
reflects the instrumental intro before the vocal and beat fully establish the 
harmonic centre.
:::

---

**🏝️ NI BORRACHO · Quevedo — Keygram**
```{r}
#| echo: false
#| warning: false
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("NI BORRACHO — Keygram")
```

::: {.callout-note}
**Key:** The keygram shows more evenly distributed brightness than DtMF, with no
dominant key throughout. There is some concentration in the upper keys, 
consistent with the track's harmonically blended character noted in the 
chromagram. The absence of a strongly dominant key across the full track 
could reflect Quevedo's intermediate Atlantic position, neither as tonally fixed
as Caribbean reggaeton nor as modally complex as Peninsular flamenco.
:::

---

**🇪🇸 Malamente · Rosalía — Keygram**
```{r}
#| echo: false
#| warning: false
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("Malamente — Keygram")
```

::: {.callout-note}
**Key:** The most distinctive keygram of the three. Clear yellow bands appear at
specific moments (at middle of song and toward the end) rather than consistently
throughout, suggesting strong but shifting tonal centres. This is characteristic
of flamenco's harmony, which does not conform to standard major/minor key 
profiles — explaining why the brightness is more fragmented and less stable than
in the other two tracks.
:::

## Timbre Features

These cepstrograms show the timbral texture over time for each track using Mel-Frequency Cepstral Coefficients (MFCCs).

**🇵🇷 Bad Bunny – DtMF (Puerto Rico)**
```{r}
#| echo: false
DTMF_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("DtMF - Bad Bunny (Puerto Rico)")
```
MFCC_02 shows the most variation with spikes throughout, indicating strong and 
consistent low-frequency energy, typical of reggeaton's heavy bass and kick drum
production.The upper MFCCs (15 to 21) are relatively uniform, suggesting a 
stable, electronically produced timbral texture with little acoustic variation. 


**🏝️ Quevedo – NI BORRACHO (Canary Islands)**
```{r}
#| echo: false
niBorracho_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("NI BORRACHO - Quevedo (Canary Islands)")
```
Just like DtMF, mfcc_02 also dominates with strong brightness. However, mfcc_07 
shows more dramatic dark patches, moments of contrast that could correspond to 
the chorus of the song and the production drops that it has. The overall texture
is more varied than Bad Bunny's, hinting at a more melodically dynamic 
production style that blends urban and pop.

**🇪🇸 Rosalía – Malamente (Spain)**
```{r}
#| echo: false
malamente_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("MALAMENTE - Rosalía (Peninsular Spain)")
```
This is the most distinctive cepstogram of the three. 
MFCC_09 shows a persistent band of negative values for 
the entire length of the track which reflects the 
timbral quality of the guitar and flamenco vocals which
occupy a different frequency than electronic production.
MFCC_02 still shows brightness but less so than the 
other two tracks.

> All three tracks share a strong mfcc_02 component, suggesting comparable 
low-frequency energy across regions. Bad Bunny and Quevedo are timbally closer 
to each other and Malamente stands out with its unique MFCC_09 pattern. 
This aligns with out hypothesis that Canarian urbano leans Caribbean rather than
Iberian.

## Temporal Features

Self-similarity matrices reveal the internal structure of each track by 
comparing every moment against every other moment. Bright diagonal stripes 
indicate repeated sections; checkerboard patterns suggest clear verse/chorus 
alternation. Each track is analysed using both chroma (harmonic) and timbre 
(textural) data.

---

### 🇵🇷 DtMF · Bad Bunny (Puerto Rico)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Chroma SSM")

DTMF_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Very dense, almost uniform bright pattern with faint diagonal 
lines which reflects reggaeton's highly repetitive harmonic loop where the same 
chord progression repeats throughout with minimal variation.
:::

::: {.callout-important}
**Timbre SSM:** Structured with clear rectangular blocks. While the harmonic content changes constantly, the timbral texture shifts between sections. Notably, the dark square visible towards the end likely corresponds to the spoken word outro where the music fades and vocals take over, creating a timbral texture distinct from the rest of the track.
:::

---

### 🏝️ NI BORRACHO · Quevedo (Canary Islands)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Chroma SSM")

niBorracho_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Similar dense pattern to Bad Bunny but with slightly more visible block structure, hinting at clearer harmonic contrast between sections. This is consistent with a more melodically varied urban-pop production style.
:::

::: {.callout-important}
**Timbre SSM:** The most striking of the three. The SSM shows a very clear large dark square in the centre surrounded by brighter areas, suggesting a distinct middle section that sounds timbrally very different from the rest. This likely corresponds to the mid-song beat change, where production shifts significantly, creating a timbral texture that contrasts sharply with both opening and closing sections. As with DtMF, a dark region towards the end likely reflects the spoken or faded-out passage.
:::

---

### 🇪🇸 Malamente · Rosalía (Peninsular Spain)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Chroma SSM")

malamente_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Shows the clearest block structure of the three: distinct sections reflecting flamenco's more varied harmonic movement compared to reggaeton's loops. Structure is visible and meaningful here.
:::

::: {.callout-important}
**Timbre SSM:** Strong diagonal stripes with clear off-diagonal blocks. Several sections return and repeat, consistent with flamenco's structured forms. Malamente has the most clearly defined musical structure of the three tracks.
:::

### Rythm and Tempo · NI BORRACHO
To extend the temporal analysis beyond structural repetition, I analysed
**NI BORRACHO** using a novelty function and two tempograms generated in
Sonic Visualiser. These visualisations highlight moments of rhythmic change
and reveal how pulse strength is distributed across different tempo levels.
```{r}
#| echo: false
ggplot(niBorracho_novelty, aes(x = TIME, y = VALUE)) +
  geom_line(color = "#00b4a2", linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title = "NI BORRACHO — Novelty Function",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Novelty"
  ) +
  atlas_theme()
```
The novelty curve shows rapid frame-level changes in the audio signal. The teal
line represents the raw novelty function. In the first seconds of **NI BORRACHO**, the novelty values are dense and high, reflecting frequent rhythmic events associated with percussion and vocal entry. Later sections show fewer spikes,suggesting more stable musical texture.

Compared to an energy-based novelty function, this spectrum-based novelty curve
captures changes in timbre and spectral structure rather than only amplitude.
For highly compressed urbano productions, this makes it more reliable for
detecting rhythmic events and therefore more useful for beat tracking.

```{r}
#| echo: false
#| warning: false
#| message: false
long_tempogram(niBorracho_act) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_y_continuous(
    transform = c("reciprocal", "reverse"),
    breaks = seq(50, 250, 50)
  ) +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title = "NI BORRACHO — Autocorrelation Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Tempo (BPM)"
  ) +
  atlas_theme()
```
The autocorrelation tempogram allowes us to observe how tempo-related structures evolve over time. Brighter horizontal bands indicate tempi with stronger periodic repetition.

In the first 30 seconds of “Ni Borracho”, several strong horizontal bands appear between approximately 70 BPM and 150 BPM, indicating a stable rhythmic structure typical of urbano and reggaeton production. The most prominent band occurs around ~90–100 BPM, which likely corresponds to the track’s underlying pulse. Additional bands at higher tempi represent harmonic multiples of the beat (double or triple tempo interpretations).

The brightness of these bands is strongest during the first 10–15 seconds, suggesting that rhythmic activity and onset density are highest in the introductory section. After this point, the energy becomes more diffuse, indicating a more stable groove where fewer new rhythmic events are introduced.

Overall, the tempogram shows that the track maintains a consistent tempo structure, with periodicities remaining relatively stable across the analysed time window.

```{r}
#| warning: false
#| message: false
#| echo: false
long_tempogram(niBorracho_dft) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30), ylim = c(40, 250)) +
  labs(
    title = "NI BORRACHO — Fourier Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Tempo (BPM)"
  ) +
  atlas_theme()
```
The Fourier tempogram shows the strength of periodicities in the novelty curve using a short-time Fourier transform. Brighter regions indicate tempi where rhythmic periodicities are strongest over time.

In the first 30 seconds of “Ni Borracho”, the most prominent band appears around 150 BPM, indicating a strong and stable rhythmic periodicity at this tempo. This likely corresponds to the perceived pulse of the track or a metrical subdivision of the main beat, which is common in urbano and reggaeton rhythms where the perceived beat may be interpreted at multiple metrical levels.

The strong horizontal structure indicates that the tempo remains stable over time, with little temporal drift. Compared with the autocorrelation tempogram, the Fourier tempogram shows fewer parallel tempo bands and highlights the dominant periodicity more clearly, making it easier to identify the principal tempo of the track.

The intensity of the band is strongest in the first 10–12 seconds, reflecting a higher density of rhythmic onsets during the introduction of the track. After this point the periodic structure remains visible but slightly weaker, suggesting a more consistent groove once the rhythm stabilizes.

## Classification of regional styles

While the previous sections analysed three representative tracks in detail,
this section scales up to the entire corpus of approximately 240 tracks.
The goal is to test whether Spotify audio features alone are sufficient to
distinguish regional styles, and if so, which features carry the most discriminating power.

A **Random Forest classifier** is trained to predict region (Puerto Rico,
Canary Islands, or Mainland Spain) from nine Spotify audio features:
danceability, energy, loudness, speechiness, acousticness,
instrumentalness, liveness, valence, and tempo. 

Random Forest was chosen because it handles correlated features naturally — energy and loudness are known to co-vary — and because its feature importance output connects directly back to the acoustic analyses in earlier sections. To reduce overfitting on the small dataset, trees are constrained with `min_n = 40`,
and `mtry` is tuned via grid search across five values. The classifier is
evaluated using **5-fold cross-validation**, with 33% accuracy representing
the random guessing baseline across three equally sized classes.

```{r}
#| echo: false
# labels for classification
artist_region <- tribble(
  ~artist_primary, ~region,
  "Bad Bunny", "Puerto Rico",
  "Rauw Alejandro", "Puerto Rico",
  "Omar Courtz", "Puerto Rico",
  "Quevedo", "Canary Islands",
  "Cruz Cafuné", "Canary Islands",
  "La Pantera", "Canary Islands",
  "Lucho RK", "Canary Islands",
  "Juseph", "Canary Islands",
  "Omar Montes", "Mainland Spain",
  "C. Tangana", "Mainland Spain",
  "ROSALÍA", "Mainland Spain",
  "rusowsky", "Mainland Spain"
)

CM_corpus <- CM_corpus |>
  mutate(
    artist_list = str_split(artist_name_s, ";\\s*")
  ) |>
  rowwise() |>
  mutate(
    region = {
      matched <- artist_region |>
        filter(artist_primary %in% artist_list)
      
      if (nrow(matched) == 0) {
        NA_character_
      } else {
        matched$region[1]
      }
    }
  ) |>
  ungroup() |>
  mutate(
    region = factor(
      region,
      levels = c("Puerto Rico", "Canary Islands", "Mainland Spain")
    )
  )

model_features <- c(
  "danceability", "energy", "loudness", "speechiness",
  "acousticness", "instrumentalness", "liveness",
  "valence", "tempo"
)

model_data <- CM_corpus |>
  select(region, all_of(model_features)) |>
  drop_na()

count(CM_corpus, region) |>
  ggplot(aes(x = region, y = n, fill = region)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, color = "#f0ecdf") +
  scale_fill_manual(values = c(
    "Puerto Rico" = "#00b4a2",
    "Canary Islands" = "#f5c842",
    "Mainland Spain" = "#d62839"
  )) +
  labs(
    title = "Corpus balance by region",
    x = NULL,
    y = "Number of tracks"
  ) +
  atlas_theme() +
  theme(legend.position = "none")
```
The corpus remains broadly balanced across the three regions, with only minor 
differences resulting from the inclusion of collaborative tracks. This confirms 
that no single region dominates the dataset, ensuring that the classification 
model is not biased toward a particular group. The slight variations also 
reflect the collaborative nature of urbano production, where artists frequently 
work across regional boundaries.

```{r}
#| echo: false
#| warning: false

# Cross-validation
set.seed(123)
rf_spec <- rand_forest(
  trees = 500,
  mtry  = tune(),
  min_n = 40
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
rf_recipe   <- recipe(region ~ ., data = model_data)
rf_workflow <- workflow() |> add_model(rf_spec) |> add_recipe(rf_recipe)
rf_grid <- grid_regular(
  mtry(range  = c(2, 6)),
  levels = 5
)
cv_folds <- vfold_cv(model_data, v = 5, strata = region)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(accuracy, kap)
)
best_params       <- select_best(rf_tune, metric = "accuracy")
rf_final_workflow <- finalize_workflow(rf_workflow, best_params)
data_split        <- initial_split(model_data, prop = 0.8, strata = region)
rf_fit            <- fit(rf_final_workflow, data = training(data_split))

show_best(rf_tune, metric = "accuracy", n = 1) |>
  bind_rows(show_best(rf_tune, metric = "kap", n = 1)) |>
  select(.metric, mean, std_err) |>
  mutate(
    .metric = recode(.metric, "accuracy" = "Accuracy", "kap" = "Cohen's κ"),
    across(c(mean, std_err), ~ round(.x, 3))
  ) |>
  rename(Metric = .metric, Mean = mean, `±SE` = std_err) |>
  gt() |>
  tab_options(
    table.background.color         = "#1f2b47",
    column_labels.background.color = "#16213e",
    table.border.top.color         = "#f5c842",
    table.border.top.width         = px(3)
  ) |>
  tab_style(
    style     = cell_text(color = "#f5c842", font = "Bebas Neue"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(color = "#f0ecdf"),
    locations = cells_body(columns = Metric)
  ) |>
  tab_style(
    style     = cell_text(color = "#00b4a2", weight = "bold"),
    locations = cells_body(columns = c(Mean, `±SE`))
  )
  
```

The tuned model achieves a cross-validated accuracy of **59.1%** 
(κ = 0.386, ±0.030), nearly double the 33% random baseline. The moderate 
κ score reflects genuine predictive signal while honestly acknowledging 
the difficulty of distinguishing regional variants of the same genre from  Spotify features.

```{r}
#| echo: false
# Training accuracy vs CV accuracy comparison
train_pred <- predict(rf_fit, training(data_split)) |>
  bind_cols(training(data_split) |> select(region))

train_acc <- accuracy(train_pred, truth = region, estimate = .pred_class)
cv_acc <- show_best(rf_tune, metric = "accuracy", n = 1)

cat("Training accuracy:", round(train_acc$.estimate, 3), "\n")
cat("CV accuracy:      ", round(cv_acc$mean, 3), "±", round(cv_acc$std_err, 3))
```

The training accuracy of **80%** against a CV accuracy of **59.1%** 
represents a considerably healthier gap than an unconstrained model. Setting `min_n = 40` forces shallower trees, successfully reducing memorisation. The remaining 20-point gap is expected given that the training 
set contains roughly 60 examples per class, some fitting to noise is 
unavoidable at this scale. What matters is that the CV accuracy is stable 
across folds (±0.030), confirming that 59% generalisation performance is 
reliable and consistent rather than a product of a lucky data split.

```{r}
#| echo: false
#| warning: false
# Confusion matrix
rf_pred    <- predict(rf_fit, testing(data_split)) |>
              bind_cols(testing(data_split) |> select(region))

conf_mat(rf_pred, truth = region, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "#12152a", high = "#f5c842") +
  labs(title = "Confusion Matrix — Regional Classifier") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    text      = element_text(color = "#f0ecdf"),
    axis.text = element_text(color = "#f0ecdf", size = 11),
    axis.title = element_text(color = "#8888aa"),
    plot.title = element_text(color = "#f5c842", size = 14, face = "bold"),
    legend.background = element_rect(fill = "#12152a", color = NA),
    legend.text  = element_text(color = "#f0ecdf"),
    legend.title = element_text(color = "#f5c842")
  )
```

The confusion matrix shows evenly distributed errors across all three 
regions, a direct consequence of the shallower trees from `min_n = 40`. 
Puerto Rico is harder to classify correctly (9/21), while Canary Islands 
and Mainland Spain remain similarly difficult (9/16 and 8/16 respectively). 
The spread of errors across all three regions confirms that no single region 
is acoustically well-separated from the others — the model is not simply 
learning one easy distinction but struggling equally with all three boundaries.

```{r}
#| echo: false
#| warning: false
# Feature importance
rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 9, aesthetics = list(fill = "#00b4a2")) +
  labs(
    title    = "Feature Importance — Random Forest",
    subtitle = "Mean decrease in node impurity (Gini)",
    x = NULL, y = "Importance"
  ) +
  coord_flip() +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid  = element_blank(),
    axis.line   = element_line(color = "#252a4a"),
    axis.ticks  = element_line(color = "#252a4a"),
    text        = element_text(color = "#f0ecdf"),
    axis.text   = element_text(color = "#f0ecdf", size = 11),
    axis.title  = element_text(color = "#8888aa"),
    plot.title  = element_text(color = "#f5c842", size = 14, face = "bold"),
    plot.subtitle = element_text(color = "#f0ecdf"),
    legend.position = "none"
  )
```
**Acousticness** remains the single most discriminating feature, followed 
closely by **loudness**, which has risen above energy compared to earlier 
model configurations. This is consistent with Malamente's distinctively 
quieter, acoustic production contrasting sharply with the heavily compressed 
loudness of reggaeton and Canarian urbano. **Tempo** ranks last, which is 
particularly striking given that tempo is a defining structural element of 
reggaeton, the dembow rhythm operates at a very specific BPM range. However, 
because Canarian urbano and mainland Spanish urban pop have both converged 
toward similar tempo zones in the streaming era, all three regions end up 
sharing the same rhythmic pace. Tempo therefore carries strong musical 
meaning within reggaeton but no *regional* discriminating power across this 
corpus, which is exactly what the tempo histogram in the Track Features tab 
showed.

```{r}
#| echo: false
#| warning: false
# ── PCA scatter ───────────────────────────────────────────────────────────
pca_baked <- recipe(region ~ ., data = model_data) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 2) |>
  prep() |>
  bake(new_data = NULL)

ggplot(pca_baked, aes(x = PC1, y = PC2, color = region)) +
  geom_point(alpha = 0.7, size = 3) +
  stat_ellipse(aes(group = region), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c(
    "Puerto Rico"    = "#00b4a2",
    "Canary Islands" = "#f5c842",
    "Mainland Spain" = "#d62839"
  )) +
  labs(
    title    = "PCA of Spotify Features by Region",
    subtitle = "Dashed ellipses = 95% confidence regions",
    x = "PC1", y = "PC2", color = NULL
  ) +
  atlas_theme() +
  theme(legend.position = "bottom")
```

The PCA scatter remains consistent with the classification results. Puerto 
Rico shows the clearest outliers along PC1, representing its most 
distinctively Caribbean tracks. Canary Islands and Mainland Spain ellipses 
overlap almost completely, explaining why these two regions generate the most 
confusion in the classifier. The heavy three-way overlap in the centre of the 
plot is the visual equivalent of the 59% accuracy ceiling for the majority 
of the corpus, regional style simply cannot be recovered from Spotify features 
alone.

---

### What the classifier tells us

The Random Forest analysis confirms the  hypothesis of this corpus 
study: 
**Canary Islands urbano does not occupy a clean intermediate position between Caribbean reggaeton and Peninsular Spanish pop: it leans Caribbean.**
The classifier struggles most to separate Canary Islands from Puerto Rico 
acoustically, while Mainland Spain is distinguishable primarily through its 
higher acousticness driven by flamenco-rooted production.

A 59% accuracy ceiling is not a failure. It is the quantitative expression 
of what the chromagram, cepstrogram, and self-similarity analyses all 
suggested qualitatively: that these three regional styles share enough 
common ground in the global streaming era that no single set of audio 
features can cleanly separate them. The features that *do* discriminate
(acousticness, loudness, energy) map directly onto the timbral differences 
identified in earlier sections, giving the classification results musical 
meaning beyond the numbers.

::: {.callout-note}
**Limitation:** With approximately 20 tracks per artist, the corpus size 
constrains classification performance. A broader corpus with more tracks 
per region would likely improve generalisation, but risks introducing 
stylistic noise if artist selection is not carefully controlled. The 
classifier should therefore be read as exploratory evidence rather than 
a definitive acoustic boundary between regions.
:::

## Contribution
:::
