---
title: "Urban Music Atlas"
format: 
  html:
    # Tabset layout chosen over format:dashboard to preserve custom CSS styling
    theme: darkly
    css: styles.css
    toc: false
    page-layout: full
---
```{r}
#| include: false
library(tidyverse)
library(compmus)
library(tidymodels)
library(ranger)
library(vip)
library(yardstick)
library(gt)

CM_corpus <- read_csv("canciones/CM_corpus.csv", show_col_types = FALSE)

CM_corpus <- CM_corpus |>
  rename_with(~ .x |>
    str_to_lower() |>
    str_replace_all("[^a-z0-9]+", "_") |>
    str_replace_all("^_|_$", ""))

DTMF <- read_csv('canciones/DTMF_chromagram.csv')
niBorracho <- read_csv('canciones/niBorracho_chromagram.csv')
malamente <- read_csv('canciones/malamente_chromagram.csv')

DTMF_timbre <- read_csv('canciones/DTMF_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_timbre <- read_csv('canciones/niBorracho_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
malamente_timbre <- read_csv('canciones/malamente_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_novelty <- read_csv("canciones/niBorracho_novelty.csv", show_col_types = FALSE)
niBorracho_act <- read_csv("canciones/niBorracho_act.csv", show_col_types = FALSE)
niBorracho_dft <- read_csv("canciones/niBorracho_dft.csv", show_col_types = FALSE)

# Shared plot theme
atlas_theme <- function() {
  theme_classic() +
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 10),
      axis.title       = element_text(color = "#8888aa", size = 11),
      plot.title       = element_text(color = "#f5c842", size = 16, face = "bold"),
      plot.subtitle    = element_text(color = "#f0ecdf", size = 11),
      plot.margin      = margin(12, 16, 12, 12),
      legend.background = element_rect(fill = "#12152a", color = NA),
      legend.key        = element_rect(fill = "#12152a", color = NA),
      legend.text       = element_text(color = "#f0ecdf"),
      legend.title      = element_text(color = "#f5c842")
    )
}

keygram_theme <- function(title_text) {
  list(
    geom_tile(color = "#12152a", linewidth = 0.3),
    scale_fill_viridis_c(guide = "none"),
    theme_classic(),
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 7),
      axis.text.y      = element_text(color = "#f0ecdf", size = 7),
      axis.title       = element_text(color = "#8888aa"),
      plot.title       = element_text(color = "#f5c842", family = "sans",
                                      size = 13, face = "bold",
                                      margin = margin(b = 10)),
      plot.margin      = margin(12, 16, 12, 12)
    ),
    labs(x = "Time (s)", y = "", title = title_text)
  )
}

# Key templates for keygrams
circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

major_key <- c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <- c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

key_templates <- tribble(
  ~name, ~template,
  "Gb:maj", circshift(major_key, 6),
  "Bb:min", circshift(minor_key, 10),
  "Db:maj", circshift(major_key, 1),
  "F:min",  circshift(minor_key, 5),
  "Ab:maj", circshift(major_key, 8),
  "C:min",  circshift(minor_key, 0),
  "Eb:maj", circshift(major_key, 3),
  "G:min",  circshift(minor_key, 7),
  "Bb:maj", circshift(major_key, 10),
  "D:min",  circshift(minor_key, 2),
  "F:maj",  circshift(major_key, 5),
  "A:min",  circshift(minor_key, 9),
  "C:maj",  circshift(major_key, 0),
  "E:min",  circshift(minor_key, 4),
  "G:maj",  circshift(major_key, 7),
  "B:min",  circshift(minor_key, 11),
  "D:maj",  circshift(major_key, 2),
  "F#:min", circshift(minor_key, 6),
  "A:maj",  circshift(major_key, 9),
  "C#:min", circshift(minor_key, 1),
  "E:maj",  circshift(major_key, 4),
  "G#:min", circshift(minor_key, 8),
  "B:maj",  circshift(major_key, 11),
  "D#:min", circshift(minor_key, 3)
)

# Small helper to safely convert tempogram column names to numeric tempo bins
long_tempogram <- function(df) {
  time_col <- names(df)[1]

  df |>
    pivot_longer(-all_of(time_col), names_to = "tempo", values_to = "value") |>
    mutate(
      tempo = as.numeric(tempo)
    ) |>
    rename(TIME = all_of(time_col))
}

# labels for regions
artist_region <- tribble(
  ~artist_primary, ~region,
  "Bad Bunny", "Puerto Rico",
  "Rauw Alejandro", "Puerto Rico",
  "Omar Courtz", "Puerto Rico",
  "Quevedo", "Canary Islands",
  "Cruz Cafuné", "Canary Islands",
  "La Pantera", "Canary Islands",
  "Lucho RK", "Canary Islands",
  "Juseph", "Canary Islands",
  "Omar Montes", "Mainland Spain",
  "C. Tangana", "Mainland Spain",
  "ROSALÍA", "Mainland Spain",
  "rusowsky", "Mainland Spain"
)

CM_corpus <- CM_corpus |>
  mutate(
    artist_list = str_split(artist_name_s, ";\\s*")
  ) |>
  rowwise() |>
  mutate(
    region = {
      matched <- artist_region |>
        filter(artist_primary %in% artist_list)
      
      if (nrow(matched) == 0) {
        NA_character_
      } else {
        matched$region[1]
      }
    }
  ) |>
  ungroup() |>
  mutate(
    region = factor(
      region,
      levels = c("Puerto Rico", "Canary Islands", "Mainland Spain")
    )
  )

region_colours <- c(
  "Puerto Rico"    = "#00b4a2",
  "Canary Islands" = "#f5c842",
  "Mainland Spain" = "#d62839"
)
```

::: hero-banner
URBAN MUSIC ATLAS
:::

Spanish-language urban music (2015–2025) across three geographic contexts.

::: panel-tabset

## Background

<iframe src="https://open.spotify.com/embed/playlist/0Z93ne5Yrdqr1hLlRqF7hq" 
width="100%" height="380" frameborder="0" 
allowtransparency="true" allow="encrypted-media"
style="border-radius: 12px;"></iframe>

### Why this corpus?

Nowadays, there is a rise in popularity of Canarian music across Spain, and with 
it, a recurring debate in music criticism and on social media, *are artists like* 
*Quevedo genuinely expressing an Atlantic identity rooted in the Canary* 
*Islands, or are they simply mimicking Puerto Rican reggaeton?* 

This portfolio takes that question seriously and tries to answer it with data 
rather than opinion.

It is also worth acknowledging that **reggaeton is culturally Puerto Rican**. 
The genre was created by Puerto Rican and broader Caribbean communities, and 
any analysis of its spread must recognise that origin without flattening it into 
a generic "Latin sound." This connection is not new, Canarian emigration to 
Cuba, Venezuela, and Puerto Rico over five centuries created a shared *español* 
*atlántico*. The goal here is not to claim equivalence between regions, but to 
measure acoustic distance, and to let that distance speak.

> *Is Canarian urbano its own sound or is it just Bad Bunny with a Spanish accent?*

---

### The corpus

The corpus covers Spanish-language urban music from **2015 to 2025** across three 
geographic regions: **Puerto Rico**, the **Canary Islands**, and **Mainland Spain**.
Approximately **240 tracks** were chosen from representative artists in each 
region, balancing reggaeton performers with hybrid urban artists. Tracks 
were retrieved via the Spotify API and selected to reflect each region's distinct
production ecosystem while keeping the corpus balanced enough for classification.

::: {.callout-note}
**Corpus assumption:** Artists were assigned to a single region based on their 
primary cultural origin, even when their music crosses regional lines. This is 
a deliberate simplification. The goal is to isolate regional signal, not to deny
that these scenes are in constant dialogue with each other.
:::

---

### My three tracks

Three tracks were submitted to the class corpus (one per region) selected for 
their **cultural rootedness** and **popularity**. Each is 
emblematic of its region's production style, making them ideal anchors for 
cross-regional comparison.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DtMF · Bad Bunny** — Puerto Rico
:::
Bad Bunny is one of the most-streamed Latin artists of all time and the most 
prominent voice in global reggaeton. **DtMF** was chosen because it is deeply 
rooted in dembow rhythm, minor-key trap production, and Boricua street 
vernacular.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands
:::
Quevedo became one of Spain's most-streamed artists with a sound heavily 
influenced by Latin Caribbean music. **NI BORRACHO** was chosen because it 
sits at the precise cultural crossroads this study investigates: rhythmically 
Caribbean, lyrically Canarian, produced in a style that borrows from reggaeton 
but is not reducible to it. It is the track that most directly embodies the 
central question of this corpus.
:::

::: {.chroma-card .sp}
::: chroma-header
🇪🇸 **Malamente · Rosalía** — Mainland Spain
:::
Rosalía is a Grammy winner who fuses contemporary production 
with deep flamenco tradition. **Malamente** was chosen as a deliberate 
counterpoint: it is the track in this corpus most acoustically distant from 
reggaeton, anchoring the Peninsular Spain end of the regional spectrum. Its 
inclusion tests whether Canarian urbano leans toward this end or away from it.
:::

:::


> *Are there measurable stylistic differences across regions?*

## Track Features

Three hypotheses guide this section, each testable against the data:

::: {.callout-note}
**H1 -> Tempo convergence:** Streaming-era production has homogenised tempo across 
all three regions, so regional origin cannot be recovered from BPM alone.

**H2 —> Danceability floor:** Danceability is uniformly high across the corpus, 
reflecting the shared dance-oriented production of all three urban scenes.

**H3 —> Canary Islands as Atlantic bridge:** The Canary Islands occupies a 
measurably intermediate acoustic position between Puerto Rico and Mainland Spain 
on at least some features, reflecting its dual Atlantic identity.
:::

---

### Tempo
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  ggplot(aes(x = tempo, fill = region)) +
  geom_histogram(binwidth = 5, color = "#0a0c14", alpha = 0.85) +
  facet_wrap(~region, ncol = 1) +
  scale_fill_manual(values = region_colours) +
  labs(
    title    = "Tempo distribution by region",
    subtitle = "All three regions share the same rhythmic window",
    x = "Tempo (BPM)", y = "Tracks"
  ) +
  atlas_theme() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "#16213e", color = NA),
        strip.text = element_text(color = "#f5c842", size = 10, face = "bold"))
```

Regional medians are nearly identical: Puerto Rico **109 BPM**, Canary Islands 
**113 BPM**, Mainland Spain **106 BPM**. All three distributions peak in the 
same 90–140 BPM window. **H1 is confirmed**, tempo carries musical meaning 
within reggaeton but no regional discriminating power across this corpus.

> Note: The cluster of Puerto Rico tracks above 170 BPM likely reflects Spotify's 
tendency to estimate dembow rhythms at double their perceived tempo.

---

### Danceability
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  ggplot(aes(x = region, y = danceability, fill = region)) +
  geom_violin(alpha = 0.75, trim = TRUE, color = NA) +
  geom_boxplot(width = 0.12, fill = "#0a0c14", color = "#f0ecdf",
               outlier.size = 1, outlier.alpha = 0.5) +
  scale_fill_manual(values = region_colours) +
  labs(
    title    = "Danceability by region",
    subtitle = "Medians: PR = 0.69 · CI = 0.68 · Spain = 0.68",
    x = NULL, y = "Danceability"
  ) +
  atlas_theme() +
  theme(legend.position = "none")
```

Median danceability is 0.691 for Puerto Rico, 0.680 for the Canary Islands, and 
0.675 for Mainland Spain: a difference of less than 0.02 across the entire 
range. **H2 is confirmed.** Danceability is a genre floor, not a regional 
signature. Urban music is dance music regardless of whether it is produced in 
San Juan, Las Palmas, or Madrid.

---

### Where the regions actually diverge
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  select(region, valence, energy, acousticness) |>
  pivot_longer(-region, names_to = "feature", values_to = "value") |>
  mutate(feature = str_to_title(feature)) |>
  ggplot(aes(x = region, y = value, fill = region)) +
  geom_violin(alpha = 0.75, trim = TRUE, color = NA) +
  geom_boxplot(width = 0.12, fill = "#0a0c14", color = "#f0ecdf",
               outlier.size = 0.8, outlier.alpha = 0.4) +
  facet_wrap(~feature, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = region_colours) +
  labs(
    title    = "Feature distributions by region",
    x = NULL, y = NULL, fill = NULL
  ) +
  atlas_theme() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    strip.background = element_rect(fill = "#16213e", color = NA),
    strip.text = element_text(color = "#f5c842", size = 10, face = "bold")
  )
```

Three clear patterns emerge:

- **Acousticness** is the sharpest regional divider. Mainland Spain's median 
(0.532) is more than double Puerto Rico's (0.235) and the Canary Islands' (0.267). 
Puerto Rico and the Canary Islands are acoustically indistinguishable here but 
Mainland Spain is.

- **Valence** tells a more nuanced story. Puerto Rico scores the *lowest* median 
valence (0.396), reflecting reggaeton's tendency toward dark, minor-key 
production. The Canary Islands (0.485) and Mainland Spain (0.530) are both 
higher. On this dimension, the Canary Islands does sit between the two, 
**partially supporting H3**.

- **Energy** shows Puerto Rico highest (0.679), with the Canary Islands and 
Mainland Spain nearly tied (0.580 and 0.589). Canarian urbano is not more 
energetic than Spanish pop, it is the acousticness and loudness, not energy, 
that separates Puerto Rico from the rest.

---

### The regional fingerprint
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  group_by(region) |>
  summarise(
    Valence      = median(valence),
    Energy       = median(energy),
    Danceability = median(danceability),
    Acousticness = median(acousticness),
    .groups = "drop"
  ) |>
  pivot_longer(-region, names_to = "feature", values_to = "value") |>
  ggplot(aes(x = feature, y = value, color = region, group = region)) +
  geom_line(linewidth = 1.2, alpha = 0.9) +
  geom_point(size = 3) +
  scale_color_manual(values = region_colours) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Regional feature fingerprints",
    subtitle = "Median Spotify features per region — parallel coordinates",
    x = NULL, y = "Median value", color = NULL
  ) +
  atlas_theme() +
  theme(legend.position = "bottom")
```

The parallel coordinates plot summarises the regional fingerprints in one view. 
The three lines converge at **Danceability** confirming H2. 
They diverge sharply at **Acousticness**, where Mainland Spain 
is higher. At **Valence**, the Canary Islands line sits between Puerto Rico 
and Spain, the clearest visual evidence for H3. The overall picture supports 
the corpus hypothesis: **these three scenes share a common urban grammar but** 
**differ in their acoustic texture**, and it is that texture, not rhythm, that 
carries regional identity.

::: {.callout-important}
**Assumption:** Regional medians are shaped by artist selection. Omar Montes 
Latin-influenced pop pushes Mainland Spain's valence upward, Rosalía's flamenco 
pushes its acousticness up. A different artist selection would change these 
medians. The patterns are real but not universal.
:::

## Loudness

Loudness in the Spotify API is measured in **decibels (dB)**, where values 
closer to 0 dB are louder and more negative values are quieter. This section 
examines how loudness varies across regions and what it reveals about production 
style in each scene.

---

### Loudness distribution by region
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  ggplot(aes(x = region, y = loudness, fill = region)) +
  geom_violin(alpha = 0.75, trim = TRUE, color = NA) +
  geom_boxplot(width = 0.12, fill = "#0a0c14", color = "#f0ecdf",
               outlier.size = 1, outlier.alpha = 0.5) +
  scale_fill_manual(values = region_colours) +
  coord_cartesian(ylim = c(-20, 0)) +
  labs(
    title    = "Loudness distribution by region",
    subtitle = "Medians: PR = −5.1 dB · CI = −7.2 dB · Spain = −6.6 dB",
    x = NULL, y = "Loudness (dB)"
  ) +
  atlas_theme() +
  theme(legend.position = "none")
```

Puerto Rico is the loudest region (median **−5.1 dB**), consistent with 
reggaeton's heavily compressed, mastered-for-streaming production. More 
striking is Puerto Rico's spread, a standard deviation of **6.7 dB** compared 
to just 2.5 dB for the Canary Islands. This reflects the stylistic range within 
the Puerto Rican selection: from heavily compressed club tracks to sparse, 
quiet album interludes. The Canary Islands and Mainland Spain are tighter 
distributions, suggesting more consistent production loudness within each scene.

---

### Loudness and energy — a production signature
```{r}
#| echo: false
#| warning: false
CM_corpus |>
  filter(!is.na(region)) |>
  ggplot(aes(x = loudness, y = energy, color = region)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~region) +
  coord_cartesian(xlim = c(-20, 0)) +
  scale_color_manual(values = region_colours) +
  labs(
    title    = "Loudness vs energy by region",
    subtitle = "Correlation: PR = 0.73 · CI = 0.85 · Spain = 0.82",
    x = "Loudness (dB)", y = "Energy", color = NULL
  ) +
  atlas_theme() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "#16213e", color = NA),
    strip.text = element_text(color = "#f5c842", size = 10, face = "bold")
  )
```

Loudness and energy are strongly correlated across all three regions 
(r = 0.73–0.85), confirming that louder tracks are consistently perceived 
as more energetic. The Canary Islands shows the tightest correlation (0.845), 
suggesting its production style is the most consistent, loud tracks are 
energetic, quiet tracks are not. Puerto Rico's looser correlation (0.734) 
reflects the same within-region diversity seen in the loudness distribution: 
some tracks are loud but surprisingly low in Spotify's energy estimate, 
likely due to heavily compressed production that reads as loud without the 
dynamic variation Spotify associates with high energy.

---

### The chosen three tracks in context
```{r}
#| echo: false
#| warning: false
my_three <- CM_corpus |>
  filter(track_name %in% c("DtMF", "NI BORRACHO", "MALAMENTE - Cap.1: Augurio"))

CM_corpus |>
  filter(!is.na(region)) |>
  ggplot(aes(x = loudness, y = energy)) +
  geom_point(aes(color = region), alpha = 0.3, size = 2) +
  geom_point(data = my_three, aes(color = region),
             size = 5, shape = 18) +
  ggrepel::geom_label_repel(
    data = my_three,
    aes(label = track_name, color = region),
    fill  = "#0a0c14",
    size  = 3.5,
    fontface = "bold",
    box.padding = 0.5
  ) +
  scale_color_manual(values = region_colours) +
  labs(
    title    = "Your three tracks vs the corpus",
    subtitle = "Diamonds = your tracks · Loudness vs energy",
    x = "Loudness (dB)", y = "Energy", color = NULL
  ) +
  atlas_theme() +
  theme(legend.position = "bottom")
```

The three focal tracks sit in revealing positions within the corpus. 
**NI BORRACHO** (−3.4 dB, energy 0.82) is one of the loudest, most energetic 
tracks in the entire corpus — sitting at the top right of the scatter, beyond 
even most Puerto Rican tracks. **MALAMENTE** (−4.8 dB, energy 0.48) is 
moderately loud but low in energy, consistent with flamenco's acoustic 
intensity without electronic compression. **DtMF** is the most striking 
outlier at −27.4 dB a very quiet track for a reggaeton artist, reflecting its 
role as a sparse, melancholic album outro rather than a club production (even if 
it always sounds in the club). This reinforces a key assumption, Bad Bunny's
catalogue spans from club productions to sparse, introspective interludes, and 
DtMF represents the latter. The outlier is not a flaw in the corpus, it is 
evidence of Puerto Rican urbano's stylistic range compared to the more 
consistent production loudness of the Canary Islands and Mainland Spain.

::: {.callout-note}
**Assumption:** Spotify's loudness measure is an integrated track-level value 
and does not capture dynamic variation within a track. A highly dynamic track 
with quiet verses and a loud chorus will return a single average value. 
Section-level loudness analysis would require audio data beyond what the 
Spotify API provides.
:::

## Chroma Features

Chromagrams map pitch-class energy over time, brighter colours indicate stronger 
presence of a given note at that moment. Keygrams compare those chroma 
features against major and minor key profiles. 
Together they answer: *what does each track sound like harmonically, and does* 
*that match what we hear?*

---

### 🇵🇷 DtMF · Bad Bunny — Puerto Rico
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
DTMF |>
  compmus_wrangle_chroma() |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text = element_text(color = "#f0ecdf"),
    axis.text = element_text(color = "#8888aa")
  )
```

DtMF is not a typical reggaeton track, it is a 
[*plena*](https://en.wikipedia.org/wiki/Plena), a Puerto Rican folk 
with choir chants and a nostalgic, mellow beat. The chromagram reflects this: 
**A♭ and E** carry the most energy, consistent with the track's key of E major. 
The brightness is relatively uniform across the entire song, suggesting a 
harmonically stable loop structure, typical of plena, the same chord progression
cycling under Bad Bunny's emotional reflections about people he misses and a 
Puerto Rico he fears losing to gentrification. The choir chants that enter 
periodically (*"debí tirar más fotos de cuando te tuve"*) likely correspond to 
the moments of slightly increased brightness across the upper pitch classes.

**🇵🇷 DtMF — Keygram**
```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 4
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = NULL, title = "DtMF — Keygram") +
  theme_minimal() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#f0ecdf", size = 7),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```

::: {.callout-note}
The keygram shows consistent brightness around **D#:min and B:maj** both 
compatible with E major, DtMF's confirmed key. The bright 
spike in the first ~25 seconds corresponds to the sparse instrumental intro 
before Bad Bunny's voice. Once the vocals and choir enter the key profile 
concentrates around D#:min and B:maj, though darker patches at roughly 100s and 
150s suggest brief harmonic departures, corresponding to the chorus sections
where the choir's call-and-response shifts the pitch energy momentarily before returning to the established tonal centre.
:::

---

### 🏝️ NI BORRACHO · Quevedo — Canary Islands
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
niBorracho |>
  compmus_wrangle_chroma() |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text = element_text(color = "#f0ecdf"),
    axis.text = element_text(color = "#8888aa")
  )
```

NI BORRACHO is a celebration of Canarian identity, a defiant anthem where 
Quevedo lists places across the islands and declares he would not leave *even* 
*if drunk* (the title translated). The chromagram shows energy spread 
across **B, Bb, A, G, F, and E** with no single dominant pitch class. 
This harmonic diffusion is characteristic of trap-urban production that blend 
chord colours. Unlike DtMF's plena loop, the chroma remains consistently active 
from the first seconds, with bright flashes across multiple pitch classes 
reflecting the track's relentless dance energy throughout. There is no dramatic 
structural contrast between sections, the production maintains the same 
harmonic density from intro to outro. The comparatively scattered chroma 
profile places this track closer to Bad Bunny than to Rosalía, despite its 
Canarian lyrical content.

**🏝️ NI BORRACHO — Keygram**
```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 4
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = NULL, title = "NI BORRACHO — Keygram") +
  theme_minimal() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#f0ecdf", size = 7),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```

::: {.callout-note}
The keygram shows almost uniformly bright values across all 24 keys throughout 
the entire track — a visual confirmation that no single key dominates. This 
is unusual and tells us something important: the key-matching algorithm 
struggles with NI BORRACHO because its harmonic language does not conform 
cleanly to standard Western major/minor profiles. The one notable exception 
is the isolated dark patch around 150 seconds, which likely corresponds to 
a brief production drop or instrumental break where the harmonic density 
temporarily reduces. The absence of a clear tonal anchor reinforces the 
Atlantic hybrid nature of the production: neither as modally fixed as 
Caribbean plena nor as tonally complex as Andalusian flamenco.
:::

---

### 🇪🇸 Malamente · Rosalía — Mainland Spain
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
malamente |>
  compmus_wrangle_chroma() |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text = element_text(color = "#f0ecdf"),
    axis.text = element_text(color = "#8888aa")
  )
```

Malamente is the opening chapter of Rosalía's *El Mal Querer*, a track 
about a bad omen, accompanied by rhythmic handclaps and organ-like synths. 
It is in **F minor** and runs just 2 minutes 29 seconds. The chromagram shows a 
strong concentration around **G, C, and F#**, reflecting the 
[Phrygian mode](https://en.wikipedia.org/wiki/Flamenco_mode) that gives flamenco 
its characteristic dark, tense quality. Notice the **clear vertical shifts** 
roughly every 30 seconds: these correspond to Rosalía's structured verse 
sections, where the handclap rhythm resets and her vocal phrases shift 
register. The brief moment of wider pitch spread near the track's centre 
(around 100 seconds) coincides with the commanding chorus where 
she sings *"malamente, mal muy mal muy mal"*. Unlike DtMF and NI BORRACHO, the 
chroma here is clearly structured rather than looped, reflecting flamenco's 
through-composed approach rather than trap's cyclic production.

**🇪🇸 Malamente — Keygram**
```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 4
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = NULL, title = "Malamente — Keygram") +
  theme_minimal() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#f0ecdf", size = 7),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```

::: {.callout-note}
The most distinctive keygram of the three. Clear bright bands appear at 
**specific structural moments** rather than consistently throughout which does 
not conform neatly to standard major/minor key profiles. The bright yellow spike 
at approximately 50 seconds corresponds to the chorus, the harmonic tension 
peaks as her voice rises before settling back into the fragmented pattern. 
The fragmentary key profile is the chromatic signature of flamenco itself: a 
modal tradition that standard Western key templates struggle to capture cleanly.
:::

---

> Quevedo's chromagram sits harmonically closer to Bad Bunny than to Rosalía. 
Both DtMF and NI BORRACHO share diffuse, loop-based chroma profiles without 
a dominant tonal anchor, while Malamente's structured harmonic movement and 
Phrygian concentration set it clearly apart. The chroma evidence supports 
the corpus hypothesis: **Canarian urbano speaks Caribbean harmonically, 
not Iberian.**

---

## Timbre Features

Cepstrograms show the timbral texture of each track over time using 
Mel-Frequency Cepstral Coefficients (MFCCs). Lower MFCCs capture broad 
spectral shape (bass, warmth, body) while higher MFCCs capture fine 
timbral detail. Bright areas indicate strong energy at that coefficient; 
dark areas indicate absence.

---

### 🇵🇷 DtMF · Bad Bunny — Puerto Rico
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
DTMF_timbre |>
  compmus_wrangle_timbre() |>
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude",
       title = "DtMF — Cepstrogram") +
  scale_fill_viridis_c() +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold"),
    legend.background = element_rect(fill = "#12152a", color = NA),
    legend.text       = element_text(color = "#f0ecdf")
  )
```

**MFCC_02** dominates throughout, indicating powerful low-frequency energy from 
the bass and kick drum that underpins DtMF's plena rhythm. The upper MFCCs 
(15–22) are uniformly dark, confirming a stable electronically-produced timbral
texture with little acoustic variation. Notably **MFCC_07** shows a persistent 
dark band across the entire track, a consistent absence of mid-range spectral 
energy that is characteristic of heavily compressed electronic production where
the mid frequencies are scooped to make room for bass and vocals.

---

### 🏝️ NI BORRACHO · Quevedo — Canary Islands
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
niBorracho_timbre |>
  compmus_wrangle_timbre() |>
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude",
       title = "NI BORRACHO — Cepstrogram") +
  scale_fill_viridis_c() +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold"),
    legend.background = element_rect(fill = "#12152a", color = NA),
    legend.text       = element_text(color = "#f0ecdf")
  )
```

Like DtMF, **MFCC_02** dominates with strong brightness, confirming 
comparable low-frequency bass energy between the Caribbean and Canarian 
productions. However **MFCC_06** shows a persistent dark band similar to 
DtMF's MFCC_07 pattern, while the mid-range coefficients (MFCC_03 to MFCC_05) 
show more activity than in DtMF. These dark patches correspond to the production
drops in NI BORRACHO, moments where Quevedo's 
vocal delivery shifts register as he moves from the verse into the energetic 
chorus listing Canarian towns. The overall timbral texture is more varied than 
DtMF's, hinting at a more melodically dynamic production style 
exactly what you would expect from a Canarian artist navigating between 
Caribbean and European influences.

---

### 🇪🇸 Malamente · Rosalía — Mainland Spain
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
malamente_timbre |>
  compmus_wrangle_timbre() |>
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude",
       title = "Malamente — Cepstrogram") +
  scale_fill_viridis_c() +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold"),
    legend.background = element_rect(fill = "#12152a", color = NA),
    legend.text       = element_text(color = "#f0ecdf")
  )
```

Simlarly to the previous visualizations, this is the most distinctive 
cepstrogram of the three. **MFCC_07 and MFCC_09** show 
a persistent dark band running the entire length of the track which 
occupy a different spectral register than electronic bass production. This 
band is absent in both DtMF and NI BORRACHO, making it the clearest timbral 
dividing line between the Spanish and Caribbean/Canarian productions. 
**MFCC_02** still shows brightness but noticeably less than the other two 
tracks, consistent with Malamente's lower loudness and energy values seen 
in the Loudness tab. The mid-range MFCCs are notably stable and uniform 
reflecting the sparse, controlled acoustic production of flamenco rather 
than the dynamic electronic layering of urban music.

---

> All three tracks share a strong MFCC_02 component, suggesting comparable 
low-frequency energy across regions. However DtMF and NI BORRACHO are 
timbally far closer to each other than either is to Malamente whose 
persistent MFCC_09 dark band is the acoustic fingerprint of flamenco 
production. **This is the timbre evidence that most directly supports the** 
**corpus hypothesis: Canarian urbano sounds Caribbean, not Iberian.**

## Temporal Features

Self-similarity matrices (SSMs) reveal the internal structure of each track 
by comparing every moment against every other moment. Bright areas indicate 
high similarity — diagonal stripes mean repeated sections, off-diagonal 
blocks mean sections that return later. Each track is analysed using both 
chroma (harmonic) and timbre (textural) data.

---

### 🇵🇷 DtMF · Bad Bunny — Puerto Rico
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 5
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "DtMF — Chroma SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )

DTMF_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "DtMF — Timbre SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```

::: {.callout-note}
**Chroma SSM:** Very dense, almost uniform bright pattern with diagonal 
lines reflecting plena's highly repetitive harmonic loop where the same 
chord progression cycles throughout with minimal variation. The uniformity 
is the structural signature of Caribbean urbano: not monotony, but groove 
through repetition.
:::

::: {.callout-important}
**Timbre SSM:** Shows a grid pattern with bright horizontal and 
vertical bands indicating that several distinct timbral sections repeat and 
return throughout the track. This is more structured than the chroma SSM 
suggests, revealing that while the harmony loops uniformly, the timbral 
texture actually alternates between sections. The darker bottom-left corner
reflects the intro section sounding timbrally distinct 
from the rest of the track.
:::

---

### 🏝️ NI BORRACHO · Quevedo — Canary Islands
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 5
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "NI BORRACHO — Chroma SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )

niBorracho_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "NI BORRACHO — Timbre SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```
::: {.callout-note}
**Chroma SSM:** A clear checkerboard pattern with distinct bright and dark 
blocks showing significantly more harmonic contrast between sections than 
DtMF's uniform loop. This confirms that NI BORRACHO has a more varied 
harmonic structure, with sections that sound noticeably different from each 
other. The verse and chorus carry distinct harmonic colours, consistent with 
Quevedo's urban-pop style.
:::

::: {.callout-important}
**Timbre SSM:** The most striking of the three. A very clear large dark square 
in the centre surrounded by brighter areas suggesting a distinct middle 
section that sounds timbrally very different from both the opening and closing. 
This corresponds to the mid-song beat change in NI BORRACHO where the 
production shifts significantly before returning and there is a spoken section. 
A darker region towards the end reflects a fade, consistent with DtMF's outro 
pattern.
:::

---

### 🇪🇸 Malamente · Rosalía — Mainland Spain
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 5
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "Malamente — Chroma SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )

malamente_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Time (s)", title = "Malamente — Timbre SSM") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold")
  )
```

::: {.callout-note}
**Chroma SSM:** The clearest block structure of the three, large bright 
off-diagonal squares confirm that specific harmonic sections return, 
particularly around 50–70s and 100–120s. This is flamenco's 
structure made visible: sections that depart and consciously return, unlike 
the continuous loops of DtMF and NI BORRACHO.
:::

::: {.callout-important}
**Timbre SSM:** Shows approximately 5 large distinct rectangular blocks,
the clearest sectional architecture of the three tracks. Each block represents 
a timbral section that sounds internally consistent but distinct from the 
others. This grid-like structure is the timbre equivalent of flamenco's verse 
forms: clearly delineated, purposefully contrasted. **Flamenco is composed,** 
**urbano is looped.**
:::

---

### Rhythm and Tempo · NI BORRACHO

To extend the temporal analysis beyond structural repetition, **NI BORRACHO** 
was analysed using a novelty function and two tempograms. These visualisations 
highlight moments of rhythmic change and reveal how pulse strength is 
distributed across different tempo levels.

```{r}
#| echo: false
#| fig-width: 10
ggplot(niBorracho_novelty, aes(x = TIME, y = VALUE)) +
  geom_line(color = "#00b4a2", linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title    = "NI BORRACHO — Novelty Function",
    subtitle = "First 30 seconds",
    x = "Time (s)", y = "Novelty"
  ) +
  atlas_theme()
```

The novelty curve shows rapid frame-level changes in the audio signal. In the 
first seconds of NI BORRACHO the novelty values are high, reaching 
peaks above 25, reflecting frequent rhythmic events associated with percussion 
and vocal entry. After approximately 15 seconds the values drop sharply, 
suggesting a more stable musical texture once the groove is established. Two 
isolated spikes reappear around 29–31 seconds, likely corresponding to a 
production change or new element entering just before the 30-second mark. 
Compared to an energy-based novelty function, this curve 
captures changes in timbre and structure rather than only amplitude,
making it more reliable for detecting rhythmic events in highly compressed 
urbano productions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
long_tempogram(niBorracho_act) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_y_continuous(
    transform = c("reciprocal", "reverse"),
    breaks = seq(50, 250, 50)
  ) +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title    = "NI BORRACHO — Autocorrelation Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)", y = "Tempo (BPM)"
  ) +
  atlas_theme()
```

The autocorrelation tempogram reveals how temporal structures evolve 
over time. The two most prominent bands appear at approximately **75 BPM** 
**and 150 BPM** the latter likely being a double-time interpretation of 
the former, consistent with urbano's dembow rhythm where the beat can be 
perceived at two metrical levels simultaneously. Crucially, all bands 
remain stable and consistent across the entire 30-second window confirming that 
NI BORRACHO maintains a locked rhythmic structure from the very first beat.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
long_tempogram(niBorracho_dft) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30), ylim = c(40, 250)) +
  labs(
    title    = "NI BORRACHO — Fourier Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)", y = "Tempo (BPM)"
  ) +
  atlas_theme()
```

The Fourier tempogram highlights the dominant periodicity more clearly than 
the autocorrelation approach. The most prominent band appears at approximately 
**150 BPM**, with a fainter band at 225 BPM representing a harmonic multiple. 
Notably there is no visible band below 100 BPM, unlike the autocorrelation 
tempogram which showed activity at ~75 BPM. This difference reflects how the 
two methods weight periodicities differently. The Fourier approach isolates 
the strongest single pulse, while autocorrelation captures all metrical levels 
simultaneously. The 150 BPM band is brightest in the first 12–15 seconds and 
fades after 20 seconds, consistent with the novelty function showing higher 
onset density in the intro before the groove stabilises.

## Classification of regional styles

While the previous sections analysed three representative tracks in detail,
this section scales up to the entire corpus of approximately 240 tracks.
The goal is to test whether Spotify audio features alone are sufficient to
distinguish regional styles, and if so, which features carry the most discriminating power.

A **Random Forest classifier** is trained to predict region (Puerto Rico,
Canary Islands, or Mainland Spain) from nine Spotify audio features:
danceability, energy, loudness, speechiness, acousticness,
instrumentalness, liveness, valence, and tempo. 

Random Forest was chosen because it handles correlated features naturally — energy and loudness are known to co-vary — and because its feature importance output connects directly back to the acoustic analyses in earlier sections. To reduce overfitting on the small dataset, trees are constrained with `min_n = 40`,
and `mtry` is tuned via grid search across five values. The classifier is
evaluated using **5-fold cross-validation**, with 33% accuracy representing
the random guessing baseline across three equally sized classes.

```{r}
#| echo: false

model_features <- c(
  "danceability", "energy", "loudness", "speechiness",
  "acousticness", "instrumentalness", "liveness",
  "valence", "tempo"
)

model_data <- CM_corpus |>
  select(region, all_of(model_features)) |>
  drop_na()

count(CM_corpus, region) |>
  ggplot(aes(x = region, y = n, fill = region)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, color = "#f0ecdf") +
  scale_fill_manual(values = c(
    "Puerto Rico" = "#00b4a2",
    "Canary Islands" = "#f5c842",
    "Mainland Spain" = "#d62839"
  )) +
  labs(
    title = "Corpus balance by region",
    x = NULL,
    y = "Number of tracks"
  ) +
  atlas_theme() +
  theme(legend.position = "none")
```
The corpus remains broadly balanced across the three regions, with only minor 
differences resulting from the inclusion of collaborative tracks. This confirms 
that no single region dominates the dataset, ensuring that the classification 
model is not biased toward a particular group. The slight variations also 
reflect the collaborative nature of urbano production, where artists frequently 
work across regional boundaries.

```{r}
#| echo: false
#| warning: false

# Cross-validation
set.seed(123)
rf_spec <- rand_forest(
  trees = 500,
  mtry  = tune(),
  min_n = 40
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
rf_recipe   <- recipe(region ~ ., data = model_data)
rf_workflow <- workflow() |> add_model(rf_spec) |> add_recipe(rf_recipe)
rf_grid <- grid_regular(
  mtry(range  = c(2, 6)),
  levels = 5
)
cv_folds <- vfold_cv(model_data, v = 5, strata = region)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(accuracy, kap)
)
best_params       <- select_best(rf_tune, metric = "accuracy")
rf_final_workflow <- finalize_workflow(rf_workflow, best_params)
data_split        <- initial_split(model_data, prop = 0.8, strata = region)
rf_fit            <- fit(rf_final_workflow, data = training(data_split))

show_best(rf_tune, metric = "accuracy", n = 1) |>
  bind_rows(show_best(rf_tune, metric = "kap", n = 1)) |>
  select(.metric, mean, std_err) |>
  mutate(
    .metric = recode(.metric, "accuracy" = "Accuracy", "kap" = "Cohen's κ"),
    across(c(mean, std_err), ~ round(.x, 3))
  ) |>
  rename(Metric = .metric, Mean = mean, `±SE` = std_err) |>
  gt() |>
  tab_options(
    table.background.color         = "#1f2b47",
    column_labels.background.color = "#16213e",
    table.border.top.color         = "#f5c842",
    table.border.top.width         = px(3)
  ) |>
  tab_style(
    style     = cell_text(color = "#f5c842", font = "Bebas Neue"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(color = "#f0ecdf"),
    locations = cells_body(columns = Metric)
  ) |>
  tab_style(
    style     = cell_text(color = "#00b4a2", weight = "bold"),
    locations = cells_body(columns = c(Mean, `±SE`))
  )
  
```

The tuned model achieves a cross-validated accuracy of **59.1%** 
(κ = 0.386, ±0.030), nearly double the 33% random baseline. The moderate 
κ score reflects genuine predictive signal while honestly acknowledging 
the difficulty of distinguishing regional variants of the same genre from  Spotify features.

```{r}
#| echo: false
# Training accuracy vs CV accuracy comparison
train_pred <- predict(rf_fit, training(data_split)) |>
  bind_cols(training(data_split) |> select(region))

train_acc <- accuracy(train_pred, truth = region, estimate = .pred_class)
cv_acc <- show_best(rf_tune, metric = "accuracy", n = 1)

cat("Training accuracy:", round(train_acc$.estimate, 3), "\n")
cat("CV accuracy:      ", round(cv_acc$mean, 3), "±", round(cv_acc$std_err, 3))
```

The training accuracy of **80%** against a CV accuracy of **59.1%** 
represents a considerably healthier gap than an unconstrained model. Setting `min_n = 40` forces shallower trees, successfully reducing memorisation. The remaining 20-point gap is expected given that the training 
set contains roughly 60 examples per class, some fitting to noise is 
unavoidable at this scale. What matters is that the CV accuracy is stable 
across folds (±0.030), confirming that 59% generalisation performance is 
reliable and consistent rather than a product of a lucky data split.

```{r}
#| echo: false
#| warning: false
# Confusion matrix
rf_pred    <- predict(rf_fit, testing(data_split)) |>
              bind_cols(testing(data_split) |> select(region))

conf_mat(rf_pred, truth = region, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "#12152a", high = "#f5c842") +
  labs(title = "Confusion Matrix — Regional Classifier") +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    text      = element_text(color = "#f0ecdf"),
    axis.text = element_text(color = "#f0ecdf", size = 11),
    axis.title = element_text(color = "#8888aa"),
    plot.title = element_text(color = "#f5c842", size = 14, face = "bold"),
    legend.background = element_rect(fill = "#12152a", color = NA),
    legend.text  = element_text(color = "#f0ecdf"),
    legend.title = element_text(color = "#f5c842")
  )
```

The confusion matrix shows evenly distributed errors across all three 
regions, a direct consequence of the shallower trees from `min_n = 40`. 
Puerto Rico is harder to classify correctly (9/21), while Canary Islands 
and Mainland Spain remain similarly difficult (9/16 and 8/16 respectively). 
The spread of errors across all three regions confirms that no single region 
is acoustically well-separated from the others — the model is not simply 
learning one easy distinction but struggling equally with all three boundaries.

```{r}
#| echo: false
#| warning: false
# Feature importance
rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 9, aesthetics = list(fill = "#00b4a2")) +
  labs(
    title    = "Feature Importance — Random Forest",
    subtitle = "Mean decrease in node impurity (Gini)",
    x = NULL, y = "Importance"
  ) +
  coord_flip() +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid  = element_blank(),
    axis.line   = element_line(color = "#252a4a"),
    axis.ticks  = element_line(color = "#252a4a"),
    text        = element_text(color = "#f0ecdf"),
    axis.text   = element_text(color = "#f0ecdf", size = 11),
    axis.title  = element_text(color = "#8888aa"),
    plot.title  = element_text(color = "#f5c842", size = 14, face = "bold"),
    plot.subtitle = element_text(color = "#f0ecdf"),
    legend.position = "none"
  )
```
**Acousticness** remains the single most discriminating feature, followed 
closely by **loudness**, which has risen above energy compared to earlier 
model configurations. This is consistent with Malamente's distinctively 
quieter, acoustic production contrasting sharply with the heavily compressed 
loudness of reggaeton and Canarian urbano. **Tempo** ranks last, which is 
particularly striking given that tempo is a defining structural element of 
reggaeton, the dembow rhythm operates at a very specific BPM range. However, 
because Canarian urbano and mainland Spanish urban pop have both converged 
toward similar tempo zones in the streaming era, all three regions end up 
sharing the same rhythmic pace. Tempo therefore carries strong musical 
meaning within reggaeton but no *regional* discriminating power across this 
corpus, which is exactly what the tempo histogram in the Track Features tab 
showed.

```{r}
#| echo: false
#| warning: false
# ── PCA scatter ───────────────────────────────────────────────────────────
pca_baked <- recipe(region ~ ., data = model_data) |>
  step_normalize(all_predictors()) |>
  step_pca(all_predictors(), num_comp = 2) |>
  prep() |>
  bake(new_data = NULL)

ggplot(pca_baked, aes(x = PC1, y = PC2, color = region)) +
  geom_point(alpha = 0.7, size = 3) +
  stat_ellipse(aes(group = region), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c(
    "Puerto Rico"    = "#00b4a2",
    "Canary Islands" = "#f5c842",
    "Mainland Spain" = "#d62839"
  )) +
  labs(
    title    = "PCA of Spotify Features by Region",
    subtitle = "Dashed ellipses = 95% confidence regions",
    x = "PC1", y = "PC2", color = NULL
  ) +
  atlas_theme() +
  theme(legend.position = "bottom")
```

The PCA scatter remains consistent with the classification results. Puerto 
Rico shows the clearest outliers along PC1, representing its most 
distinctively Caribbean tracks. Canary Islands and Mainland Spain ellipses 
overlap almost completely, explaining why these two regions generate the most 
confusion in the classifier. The heavy three-way overlap in the centre of the 
plot is the visual equivalent of the 59% accuracy ceiling for the majority 
of the corpus, regional style simply cannot be recovered from Spotify features 
alone.

---

### What the classifier tells us

The Random Forest analysis confirms the  hypothesis of this corpus 
study: 
**Canary Islands urbano does not occupy a clean intermediate position between Caribbean reggaeton and Peninsular Spanish pop: it leans Caribbean.**
The classifier struggles most to separate Canary Islands from Puerto Rico 
acoustically, while Mainland Spain is distinguishable primarily through its 
higher acousticness driven by flamenco-rooted production.

A 59% accuracy ceiling is not a failure. It is the quantitative expression 
of what the chromagram, cepstrogram, and self-similarity analyses all 
suggested qualitatively: that these three regional styles share enough 
common ground in the global streaming era that no single set of audio 
features can cleanly separate them. The features that *do* discriminate
(acousticness, loudness, energy) map directly onto the timbral differences 
identified in earlier sections, giving the classification results musical 
meaning beyond the numbers.

::: {.callout-note}
**Limitation:** With approximately 20 tracks per artist, the corpus size 
constrains classification performance. A broader corpus with more tracks 
per region would likely improve generalisation, but risks introducing 
stylistic noise if artist selection is not carefully controlled. The 
classifier should therefore be read as exploratory evidence rather than 
a definitive acoustic boundary between regions.
:::

## Contribution
:::
