---
title: "Urban Music Atlas"
format: 
  html:
    theme: darkly
    css: styles.css
    toc: false
    page-layout: full
---
```{r}
#| include: false
library(tidyverse)
library(compmus)

DTMF <- read_csv('canciones/DTMF_chromagram.csv')
niBorracho <- read_csv('canciones/niBorracho_chromagram.csv')
malamente <- read_csv('canciones/malamente_chromagram.csv')

DTMF_timbre <- read_csv('canciones/DTMF_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_timbre <- read_csv('canciones/niBorracho_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
malamente_timbre <- read_csv('canciones/malamente_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
```

## Welcome {.unnumbered}

Spanish-language urban music (2015–2025) across three geographic contexts.

::: panel-tabset
## Background
This corpus examines Spanish-language urban music (2015–2025) across three 
geographic contexts: Puerto Rico, the Canary Islands, and mainland Spain. 
Puerto Rico represents the birthplace of reggaeton and a central node of 
Caribbean urban production. Mainland Spain represents a European market in which
urban styles have hybridised with pop, flamenco, and experimental influences. 
The Canary Islands occupy a unique Atlantic position: politically Spanish yet 
historically and culturally connected to Latin America and geographically closer
to Africa.

The corpus includes approximately 240 tracks drawn from representative artists 
in each region, balancing reggaeton-focused performers with hybrid urban 
artists. By analysing Spotify track-level audio features, I want to investigate
whether measurable stylistic differences reflect proximity to the Caribbean. In 
particular, whether Canary Islands urbano occupies an intermediate position 
between Caribbean reggaeton and mainland Spanish urban pop.

Rather than assuming genre boundaries, the study explores how urban music 
ecosystems diverge or converge across space within the global streaming era.

## Track Features

## Chroma Features

These three chromagrams represent the pitch-class energy over time for three 
culturally rooted tracks from each region.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DTMF · Bad Bunny** — Puerto Rico
:::
```{r}
#| echo: false
DTMF |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Shows prominent activity around A♭ and E, common in reggaeton production rooted 
in Puerto Rico.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands
:::
```{r}
#| echo: false
niBorracho |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Energy evenly spread across A, G, F, and E — no single dominant tonal centre, 
suggesting a blend of Iberian and Latin American influences.
:::

::: {.chroma-card .sp}
::: chroma-header
🇪🇸 **MALAMENTE · Rosalía** — Mainland Spain
:::
```{r}
#| echo: false
malamente |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Strong concentration around G, C, and F — reflecting the harmonic language deeply embedded in flamenco tradition.
:::

:::

> Quevedo's chromagram appears harmonically closer to Bad Bunny than to Rosalía — hinting that Canarian music carries a stronger Latin Caribbean influence than a Peninsular Spanish one.

## Timbre Features

These cepstrograms show the timbral texture over time for each track using Mel-Frequency Cepstral Coefficients (MFCCs).

**🇵🇷 Bad Bunny – DtMF (Puerto Rico)**
```{r}
#| echo: false
DTMF_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("DtMF - Bad Bunny (Puerto Rico)")
```
MFCC_02 shows the most variation with spikes throughout, indicating strong and 
consistent low-frequency energy, typical of reggeaton's heavy bass and kick drum
production.The upper MFCCs (15 to 21) are relatively uniform, suggesting a 
stable, electronically produced timbral texture with little acoustic variation. 


**🏝️ Quevedo – NI BORRACHO (Canary Islands)**
```{r}
#| echo: false
niBorracho_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("NI BORRACHO - Quevedo (Canary Islands)")
```
Just like DtMF, mfcc_02 also dominates with strong brightness. However, mfcc_07 
shows more dramatic dark patches, moments of contrast that could correspond to 
the chorus of the song and the production drops that it has. The overall texture
is more varied than Bad Bunny's, hinting at a more melodically dynamic 
production style that blends urban and pop.

**🇪🇸 Rosalía – Malamente (Spain)**
```{r}
#| echo: false
malamente_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("MALAMENTE - Rosalía (Peninsular Spain)")
```
This is the most distinctive cepstogram of the three. 
MFCC_09 shows a persistent band of negative values for 
the entire length of the track which reflects the 
timbral quality of the guitar and flamenco vocals which
occupy a different frequency than electronic production.
MFCC_02 still shows brightness but less so than the 
other two tracks.

> All three tracks share a strong mfcc_02 component, suggesting comparable 
low-frequency energy across regions. Bad Bunny and Quevedo are timbally closer 
to each other and Malamente stands out with its unique MFCC_09 pattern. 
This aligns with out hypothesis that Canarian urbano leans Caribbean rather than
Iberian.

## Temporal Features

Self-similarity matrices reveal the internal structure of each track by 
comparing every moment against every other moment. Bright diagonal stripes 
indicate repeated sections; checkerboard patterns suggest clear verse/chorus 
alternation. Each track is analysed using both chroma (harmonic) and timbre 
(textural) data.

---

### 🇵🇷 DtMF · Bad Bunny (Puerto Rico)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Chroma SSM")

DTMF_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Timbre SSM")
```

---

### 🏝️ NI BORRACHO · Quevedo (Canary Islands)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Chroma SSM")

niBorracho_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Timbre SSM")
```

---

### 🇪🇸 Malamente · Rosalía (Peninsular Spain)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Chroma SSM")

malamente_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Timbre SSM")
```

## Clustering

## Contribution
:::
