---
title: "Urban Music Atlas"
format: 
  html:
    theme: darkly
    css: styles.css
    toc: false
    page-layout: full
---
```{r}
#| include: false
library(tidyverse)
library(compmus)

CM_corpus <- read_csv("canciones/CM_corpus.csv", show_col_types = FALSE)

CM_corpus <- CM_corpus |>
  rename_with(~ .x |>
    str_to_lower() |>
    str_replace_all("[^a-z0-9]+", "_") |>
    str_replace_all("^_|_$", ""))

DTMF <- read_csv('canciones/DTMF_chromagram.csv')
niBorracho <- read_csv('canciones/niBorracho_chromagram.csv')
malamente <- read_csv('canciones/malamente_chromagram.csv')

DTMF_timbre <- read_csv('canciones/DTMF_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_timbre <- read_csv('canciones/niBorracho_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
malamente_timbre <- read_csv('canciones/malamente_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:22))),
                         skip = 1)
niBorracho_novelty <- read_csv("canciones/niBorracho_novelty.csv", show_col_types = FALSE)
niBorracho_act <- read_csv("canciones/niBorracho_act.csv", show_col_types = FALSE)
niBorracho_dft <- read_csv("canciones/niBorracho_dft.csv", show_col_types = FALSE)
# Shared plot theme
atlas_theme <- function() {
  theme_classic() +
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 10),
      axis.title       = element_text(color = "#8888aa", size = 11),
      plot.title       = element_text(color = "#f5c842", size = 16, face = "bold"),
      plot.subtitle    = element_text(color = "#f0ecdf", size = 11),
      plot.margin      = margin(12, 16, 12, 12),
      legend.background = element_rect(fill = "#12152a", color = NA),
      legend.key        = element_rect(fill = "#12152a", color = NA),
      legend.text       = element_text(color = "#f0ecdf"),
      legend.title      = element_text(color = "#f5c842")
    )
}

# Small helper to safely convert tempogram column names to numeric tempo bins
long_tempogram <- function(df) {
  time_col <- names(df)[1]

  df |>
    pivot_longer(-all_of(time_col), names_to = "tempo", values_to = "value") |>
    mutate(
      tempo = as.numeric(tempo)
    ) |>
    rename(TIME = all_of(time_col))
}
```

::: hero-banner
URBAN MUSIC ATLAS
:::

Spanish-language urban music (2015–2025) across three geographic contexts.

::: panel-tabset
## Background

<iframe src="https://open.spotify.com/embed/playlist/0Z93ne5Yrdqr1hLlRqF7hq" 
width="100%" height="380" frameborder="0" 
allowtransparency="true" allow="encrypted-media"
style="border-radius: 12px;"></iframe>

This corpus examines Spanish-language urban music (2015–2025) across three 
geographic contexts: Puerto Rico, the Canary Islands, and mainland Spain. 
Puerto Rico represents the birthplace of reggaeton and a central node of 
Caribbean urban production. Mainland Spain represents a European market in which
urban styles have hybridised with pop, flamenco, and experimental influences. 
The Canary Islands occupy a unique Atlantic position: politically Spanish yet 
historically and culturally connected to Latin America and geographically closer
to Africa.

The corpus includes approximately 240 tracks drawn from representative artists 
in each region, balancing reggaeton-focused performers with hybrid urban 
artists. By analysing Spotify track-level audio features, I want to investigate
whether measurable stylistic differences reflect proximity to the Caribbean. In 
particular, whether Canary Islands urbano occupies an intermediate position 
between Caribbean reggaeton and mainland Spanish urban pop.

Rather than assuming genre boundaries, the study explores how urban music 
ecosystems diverge or converge across space within the global streaming era.

## Track Features


## Chroma Features

These three chromagrams represent the pitch-class energy over time for three 
culturally rooted tracks from each region.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DTMF · Bad Bunny** — Puerto Rico

:::
```{r}
#| echo: false
DTMF |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Shows prominent activity around A♭ and E, common in reggaeton production rooted 
in Puerto Rico.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands

:::
```{r}
#| echo: false
niBorracho |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Energy evenly spread across A, G, F, and E — no single dominant tonal centre, 
suggesting a blend of Iberian and Latin American influences.
:::

::: {.chroma-card .sp}
::: chroma-header
🇪🇸 **MALAMENTE · Rosalía** — Mainland Spain

:::
```{r}
#| echo: false
malamente |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Strong concentration around G, C, and F — reflecting the harmonic language deeply embedded in flamenco tradition.
:::

:::

> Quevedo's chromagram appears harmonically closer to Bad Bunny than to Rosalía — hinting that Canarian music carries a stronger Latin Caribbean influence than a Peninsular Spanish one.

---

### Key Estimation
```{r}
#| echo: false
#| warning: false
#| include: false
circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

major_key <- c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <- c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

key_templates <- tribble(
  ~name, ~template,
  "Gb:maj", circshift(major_key, 6),
  "Bb:min", circshift(minor_key, 10),
  "Db:maj", circshift(major_key, 1),
  "F:min",  circshift(minor_key, 5),
  "Ab:maj", circshift(major_key, 8),
  "C:min",  circshift(minor_key, 0),
  "Eb:maj", circshift(major_key, 3),
  "G:min",  circshift(minor_key, 7),
  "Bb:maj", circshift(major_key, 10),
  "D:min",  circshift(minor_key, 2),
  "F:maj",  circshift(major_key, 5),
  "A:min",  circshift(minor_key, 9),
  "C:maj",  circshift(major_key, 0),
  "E:min",  circshift(minor_key, 4),
  "G:maj",  circshift(major_key, 7),
  "B:min",  circshift(minor_key, 11),
  "D:maj",  circshift(major_key, 2),
  "F#:min", circshift(minor_key, 6),
  "A:maj",  circshift(major_key, 9),
  "C#:min", circshift(minor_key, 1),
  "E:maj",  circshift(major_key, 4),
  "G#:min", circshift(minor_key, 8),
  "B:maj",  circshift(major_key, 11),
  "D#:min", circshift(minor_key, 3)
)
```

These keygrams compare the chroma features of each track against major and minor
key profiles, revealing which key the music gravitates towards over time. 
Brighter areas indicate a stronger match to that key.

```{r}
#| echo: false
#| warning: false

# Shared theme helper — define once, reuse
keygram_theme <- function(title_text) {
  list(
    geom_tile(color = "#12152a", linewidth = 0.3),
    scale_fill_gradientn(
      colors = c("#d62839", "#0a0c14", "#00b4a2", "#f5c842"),
      guide  = "none"
    ),
    theme_classic(),
    theme(
      plot.background  = element_rect(fill = "#0a0c14", color = NA),
      panel.background = element_rect(fill = "#12152a", color = NA),
      panel.grid       = element_blank(),
      axis.line        = element_line(color = "#252a4a"),
      axis.ticks       = element_line(color = "#252a4a"),
      text             = element_text(color = "#f0ecdf", family = "sans"),
      axis.text        = element_text(color = "#8888aa", size = 7),
      axis.text.y      = element_text(color = "#f0ecdf", size = 7),
      axis.title       = element_text(color = "#8888aa"),
      plot.title       = element_text(color = "#f5c842", family = "sans",
                                      size = 13, face = "bold",
                                      margin = margin(b = 10)),
      plot.margin      = margin(12, 16, 12, 12)
    ),
    labs(x = "Time (s)", y = "", title = title_text)
  )
}
```

**🇵🇷 DtMF · Bad Bunny — Keygram**
```{r}
#| echo: false
#| warning: false
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("DtMF — Keygram")
```

::: {.callout-note}
**Key:** The keygram shows consistent brightness concentrated around D#:min and 
B:maj throughout the track, suggesting DtMF gravitates towards these keys. 
This is consistent with reggaeton's tendency to use minor keys with a dark, 
intense character. The bright yellow spike in the first 25 seconds likely 
reflects the instrumental intro before the vocal and beat fully establish the 
harmonic centre.
:::

---

**🏝️ NI BORRACHO · Quevedo — Keygram**
```{r}
#| echo: false
#| warning: false
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("NI BORRACHO — Keygram")
```

::: {.callout-note}
**Key:** The keygram shows more evenly distributed brightness than DtMF, with no
dominant key throughout. There is some concentration in the upper keys, 
consistent with the track's harmonically blended character noted in the 
chromagram. The absence of a strongly dominant key across the full track 
could reflect Quevedo's intermediate Atlantic position, neither as tonally fixed
as Caribbean reggaeton nor as modally complex as Peninsular flamenco.
:::

---

**🇪🇸 Malamente · Rosalía — Keygram**
```{r}
#| echo: false
#| warning: false
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, method = "euclidean", norm = "manhattan") |>
  ggplot(aes(x = start + duration/2, width = 50*duration, y = name, fill = d)) +
  keygram_theme("Malamente — Keygram")
```

::: {.callout-note}
**Key:** The most distinctive keygram of the three. Clear yellow bands appear at
specific moments (at middle of song and toward the end) rather than consistently
throughout, suggesting strong but shifting tonal centres. This is characteristic
of flamenco's harmony, which does not conform to standard major/minor key 
profiles — explaining why the brightness is more fragmented and less stable than
in the other two tracks.
:::

## Timbre Features

These cepstrograms show the timbral texture over time for each track using Mel-Frequency Cepstral Coefficients (MFCCs).

**🇵🇷 Bad Bunny – DtMF (Puerto Rico)**
```{r}
#| echo: false
DTMF_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("DtMF - Bad Bunny (Puerto Rico)")
```
MFCC_02 shows the most variation with spikes throughout, indicating strong and 
consistent low-frequency energy, typical of reggeaton's heavy bass and kick drum
production.The upper MFCCs (15 to 21) are relatively uniform, suggesting a 
stable, electronically produced timbral texture with little acoustic variation. 


**🏝️ Quevedo – NI BORRACHO (Canary Islands)**
```{r}
#| echo: false
niBorracho_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("NI BORRACHO - Quevedo (Canary Islands)")
```
Just like DtMF, mfcc_02 also dominates with strong brightness. However, mfcc_07 
shows more dramatic dark patches, moments of contrast that could correspond to 
the chorus of the song and the production drops that it has. The overall texture
is more varied than Bad Bunny's, hinting at a more melodically dynamic 
production style that blends urban and pop.

**🇪🇸 Rosalía – Malamente (Spain)**
```{r}
#| echo: false
malamente_timbre |>
  compmus_wrangle_timbre() |> 
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  scale_fill_viridis_c() +                              
  theme_classic() +
  ggtitle("MALAMENTE - Rosalía (Peninsular Spain)")
```
This is the most distinctive cepstogram of the three. 
MFCC_09 shows a persistent band of negative values for 
the entire length of the track which reflects the 
timbral quality of the guitar and flamenco vocals which
occupy a different frequency than electronic production.
MFCC_02 still shows brightness but less so than the 
other two tracks.

> All three tracks share a strong mfcc_02 component, suggesting comparable 
low-frequency energy across regions. Bad Bunny and Quevedo are timbally closer 
to each other and Malamente stands out with its unique MFCC_09 pattern. 
This aligns with out hypothesis that Canarian urbano leans Caribbean rather than
Iberian.

## Temporal Features

Self-similarity matrices reveal the internal structure of each track by 
comparing every moment against every other moment. Bright diagonal stripes 
indicate repeated sections; checkerboard patterns suggest clear verse/chorus 
alternation. Each track is analysed using both chroma (harmonic) and timbre 
(textural) data.

---

### 🇵🇷 DtMF · Bad Bunny (Puerto Rico)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
DTMF |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Chroma SSM")

DTMF_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "DtMF — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Very dense, almost uniform bright pattern with faint diagonal 
lines which reflects reggaeton's highly repetitive harmonic loop where the same 
chord progression repeats throughout with minimal variation.
:::

::: {.callout-important}
**Timbre SSM:** Structured with clear rectangular blocks. While the harmonic content changes constantly, the timbral texture shifts between sections. Notably, the dark square visible towards the end likely corresponds to the spoken word outro where the music fades and vocals take over, creating a timbral texture distinct from the rest of the track.
:::

---

### 🏝️ NI BORRACHO · Quevedo (Canary Islands)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
niBorracho |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Chroma SSM")

niBorracho_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "NI BORRACHO — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Similar dense pattern to Bad Bunny but with slightly more visible block structure, hinting at clearer harmonic contrast between sections. This is consistent with a more melodically varied urban-pop production style.
:::

::: {.callout-important}
**Timbre SSM:** The most striking of the three. The SSM shows a very clear large dark square in the centre surrounded by brighter areas, suggesting a distinct middle section that sounds timbrally very different from the rest. This likely corresponds to the mid-song beat change, where production shifts significantly, creating a timbral texture that contrasts sharply with both opening and closing sections. As with DtMF, a dark region towards the end likely reflects the spoken or faded-out passage.
:::

---

### 🇪🇸 Malamente · Rosalía (Peninsular Spain)
```{r}
#| echo: false
#| warning: false
#| layout-ncol: 2
malamente |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(pitches, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Chroma SSM")

malamente_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  ggplot(aes(x = xstart + xduration/2, width = 50*xduration,
             y = ystart + yduration/2, height = 50*yduration, fill = d)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Malamente — Timbre SSM")
```

::: {.callout-note}
**Chroma SSM:** Shows the clearest block structure of the three: distinct sections reflecting flamenco's more varied harmonic movement compared to reggaeton's loops. Structure is visible and meaningful here.
:::

::: {.callout-important}
**Timbre SSM:** Strong diagonal stripes with clear off-diagonal blocks. Several sections return and repeat, consistent with flamenco's structured forms. Malamente has the most clearly defined musical structure of the three tracks.
:::

### Rythm and Tempo · NI BORRACHO
To extend the temporal analysis beyond structural repetition, I analysed
**NI BORRACHO** using a novelty function and two tempograms generated in
Sonic Visualiser. These visualisations highlight moments of rhythmic change
and reveal how pulse strength is distributed across different tempo levels.
```{r}
#| echo: false
ggplot(niBorracho_novelty, aes(x = TIME, y = VALUE)) +
  geom_line(color = "#00b4a2", linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title = "NI BORRACHO — Novelty Function",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Novelty"
  ) +
  atlas_theme()
```
The novelty curve shows rapid frame-level changes in the audio signal. The teal
line represents the raw novelty function. In the first seconds of **NI BORRACHO**, the novelty values are dense and high, reflecting frequent rhythmic events associated with percussion and vocal entry. Later sections show fewer spikes,suggesting more stable musical texture.

Compared to an energy-based novelty function, this spectrum-based novelty curve
captures changes in timbre and spectral structure rather than only amplitude.
For highly compressed urbano productions, this makes it more reliable for
detecting rhythmic events and therefore more useful for beat tracking.

```{r}
#| echo: false
#| warning: false
#| message: false
long_tempogram(niBorracho_act) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_y_continuous(
    transform = c("reciprocal", "reverse"),
    breaks = seq(50, 250, 50)
  ) +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30)) +
  labs(
    title = "NI BORRACHO — Autocorrelation Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Tempo (BPM)"
  ) +
  atlas_theme()
```
The autocorrelation tempogram allowes us to observe how tempo-related structures evolve over time. Brighter horizontal bands indicate tempi with stronger periodic repetition.

In the first 30 seconds of “Ni Borracho”, several strong horizontal bands appear between approximately 70 BPM and 150 BPM, indicating a stable rhythmic structure typical of urbano and reggaeton production. The most prominent band occurs around ~90–100 BPM, which likely corresponds to the track’s underlying pulse. Additional bands at higher tempi represent harmonic multiples of the beat (double or triple tempo interpretations).

The brightness of these bands is strongest during the first 10–15 seconds, suggesting that rhythmic activity and onset density are highest in the introductory section. After this point, the energy becomes more diffuse, indicating a more stable groove where fewer new rhythmic events are introduced.

Overall, the tempogram shows that the track maintains a consistent tempo structure, with periodicities remaining relatively stable across the analysed time window.

```{r}
#| warning: false
#| message: false
#| echo: false
long_tempogram(niBorracho_dft) |>
  ggplot(aes(x = TIME, y = tempo, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(option = "C", guide = "none") +
  coord_cartesian(xlim = c(0, 30), ylim = c(40, 250)) +
  labs(
    title = "NI BORRACHO — Fourier Tempogram",
    subtitle = "First 30 seconds",
    x = "Time (s)",
    y = "Tempo (BPM)"
  ) +
  atlas_theme()
```
The Fourier tempogram shows the strength of periodicities in the novelty curve using a short-time Fourier transform. Brighter regions indicate tempi where rhythmic periodicities are strongest over time.

In the first 30 seconds of “Ni Borracho”, the most prominent band appears around 150 BPM, indicating a strong and stable rhythmic periodicity at this tempo. This likely corresponds to the perceived pulse of the track or a metrical subdivision of the main beat, which is common in urbano and reggaeton rhythms where the perceived beat may be interpreted at multiple metrical levels.

The strong horizontal structure indicates that the tempo remains stable over time, with little temporal drift. Compared with the autocorrelation tempogram, the Fourier tempogram shows fewer parallel tempo bands and highlights the dominant periodicity more clearly, making it easier to identify the principal tempo of the track.

The intensity of the band is strongest in the first 10–12 seconds, reflecting a higher density of rhythmic onsets during the introduction of the track. After this point the periodic structure remains visible but slightly weaker, suggesting a more consistent groove once the rhythm stabilizes.

## Clustering
While the previous sections have analysed three representative tracks in detail, this section looks at the entire corpus of approximately 240. 

In particular tempo provides a useful rythmic dimension for comparing regional scenes.
```{r}
#| echo: false

ggplot(CM_corpus, aes(x = tempo)) +
  geom_histogram(
    binwidth = 5,
    fill = "#00b4a2",
    color = "#0a0c14"
  ) +
  labs(
    title = "Tempo distribution of the corpus",
    subtitle = "Spotify tempo estimates for the corpus",
    x = "Tempo (BPM)",
    y = "Number of tracks"
  ) +
  theme_classic() +
  theme(
    plot.background  = element_rect(fill = "#0a0c14", color = NA),
    panel.background = element_rect(fill = "#12152a", color = NA),
    panel.grid       = element_blank(),
    axis.line        = element_line(color = "#252a4a"),
    axis.ticks       = element_line(color = "#252a4a"),
    text             = element_text(color = "#f0ecdf"),
    axis.text        = element_text(color = "#8888aa"),
    axis.title       = element_text(color = "#8888aa"),
    plot.title       = element_text(color = "#f5c842", size = 13, face = "bold"),
    plot.subtitle    = element_text(color = "#f0ecdf")
  )
```
The histogram shows that most tracks cluster in a fairly tight rhythmic zone typical
of contemporary urbano and reggaeton. Much of the playlist lies roughly between
90 and 140 BPM, which is consistent with dembow-based rhythmic structures and
dance-oriented production. At the same time, the spread outside this central range
suggests hybridisation: some tracks slow down into heavier groove-based zones,
while others move upward toward faster pop-electronic tempos.

## Contribution
:::
