---
title: "Urban Music Atlas"
format: 
  html:
    theme: darkly
    css: styles.css
    toc: false
---
```{r}
#| include: false
library(tidyverse)
library(compmus)

DTMF <- read_csv('canciones/DTMF_chromagram.csv')
niBorracho <- read_csv('canciones/niBorracho_chromagram.csv')
malamente <- read_csv('canciones/malamente_chromagram.csv')

DTMF_timbre <- read_csv('canciones/DTMF_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:20))))

niBorracho_timbre <- read_csv('canciones/niBorracho_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:20))))

malamente_timbre <- read_csv('canciones/malamente_timbre.csv', 
                         col_names = c("TIME", paste0("mfcc_", sprintf("%02d", 1:20))))
```

## Welcome {.unnumbered}

Spanish-language urban music (2015–2025) across three geographic contexts.

::: panel-tabset
## Background
This corpus examines Spanish-language urban music (2015–2025) across three geographic contexts: Puerto Rico, the Canary Islands, and mainland Spain. Puerto Rico represents the birthplace of reggaeton and a central node of Caribbean urban production. Mainland Spain represents a European market in which urban styles have hybridised with pop, flamenco, and experimental influences. The Canary Islands occupy a unique Atlantic position: politically Spanish yet historically and culturally connected to Latin America and geographically closer to Africa.

The corpus includes approximately 240 tracks drawn from representative artists in each region, balancing reggaeton-focused performers with hybrid urban artists. By analysing Spotify track-level audio features, I want to investigate whether measurable stylistic differences reflect proximity to the Caribbean. In particular, whether Canary Islands urbano occupies an intermediate position between Caribbean reggaeton and mainland Spanish urban pop.

Rather than assuming genre boundaries, the study explores how urban music ecosystems diverge or converge across space within the global streaming era.

## Track Features

## Chroma Features

These three chromagrams represent the pitch-class energy over time for three culturally rooted tracks from each region.

::: chroma-grid

::: {.chroma-card .pr}
::: chroma-header
🇵🇷 **DTMF · Bad Bunny** — Puerto Rico
:::
```{r}
#| echo: false
DTMF |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Shows prominent activity around A♭ and E, common in reggaeton production rooted in Puerto Rico.
:::

::: {.chroma-card .ci}
::: chroma-header
🏝️ **NI BORRACHO · Quevedo** — Canary Islands
:::
```{r}
#| echo: false
niBorracho |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Energy evenly spread across A, G, F, and E — no single dominant tonal centre, suggesting a blend of Iberian and Latin American influences.
:::

::: {.chroma-card .sp}
::: chroma-header
🇪🇸 **MALAMENTE · Rosalía** — Mainland Spain
:::
```{r}
#| echo: false
malamente |>
  compmus_wrangle_chroma() |> 
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) |>
  compmus_gather_chroma() |> 
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()
```
Strong concentration around G, C, and F — reflecting the harmonic language deeply embedded in flamenco tradition.
:::

:::

> Quevedo's chromagram appears harmonically closer to Bad Bunny than to Rosalía — hinting that Canarian music carries a stronger Latin Caribbean influence than a Peninsular Spanish one.

## Timbre Features
```{r}
#| echo: false

```

```{r}
#| echo: false

```

```{r}
#| echo: false

```


## Clustering

## Contribution
:::
